Epidermal Resident Memory T Cell Fitness Requires Antigen Encounter in the Skin

Version 1

Abstract

Abstract

CD8+ tissue resident memory T cells (TRM) develop from effectors that seed peripheral tissues where they persist providing defense against subsequent challenges. TRM persistence requires autocrine TGFβ transactivated by integrins expressed on keratinocytes. TRM precursors that encounter antigen in the epidermis during development outcompete bystander TRM for TGFβ resulting in enhanced persistence. ScRNA-seq analysis of epidermal TRM revealed that local antigen experience in the skin resulted in an enhanced differentiation signature in comparison with bystanders. Upon recall, TRM displayed greater proliferation dictated by affinity of antigen experienced during epidermal development. Finally, local antigen experienced TRM differentially expressed TGFβRIII, which increases avidity of the TGFβRI/II receptor complex for TGFβ. Selective ablation of Tgfbr3 reduced local antigen experienced TRM capacity to persist, rendering them phenotypically like bystander TRM. Thus, antigen driven TCR signaling in the epidermis during TRM differentiation results in a lower TGFβ requirement for persistence and increased proliferative capacity that together enhance epidermal TRM fitness.

Introduction

Tissue resident CD8 memory T cells (TRM) are a highly abundant, non-circulating, long-lived subset of memory T cells that play an important role in protecting against re-infections [1, 2]. TRM also provide immunosurveillance against neoplasia and are thought to be pathogenic in some autoimmune diseases [3–9]. In the skin, following infection with vaccinia virus (VV) or herpes simplex virus, TRM develop from CD8+ T cell effectors (TEFF) that expand following priming in the lymph node and are recruited into the skin by inflammatory signals where they preferentially reside in the epidermis [1, 10–15]. Unlike TRM in some tissues, re-encounter with cognate antigen in the skin is not required for TRM differentiation [16–19]. Epidermal TRM that encounter cognate antigen in the skin or bystander TRM that do not encounter antigen both develop with equal efficiency and both express similar levels of the canonical TRM markers CD103 and CD69[20, 21].

The cytokine transforming growth factor β (TGFβ) is required for cutaneous TRM development at multiple stages. TGFβ signaling in the lymph node during steady state epigenetically preconditions naïve CD8 T cells allowing for later TRM differentiation [22]. TEFF recruited into skin require TGFβ signaling for entry into the epidermis and differentiation into TRM [23, 24]. Once epidermal TRM have differentiated, these cells continue to require TGFβ signaling to retain epidermal residence. TGFβ is produced bound to the latency associate peptide that prevents bioactivity until the complex is activated which in the epidermis is mediated exclusively by activation via the integrins αvβ6 and αvβ8 expressed by keratinocytes [25–28]. Epidermal persistence of TRM depends on autocrine TGFβ which is transactivated by the integrins ⍺vβ6 and ⍺vβ8[21, 29]. Inducible ablation in TRM of a required component of the TGFβ receptor (TGFβRII) or TGFβ1 as well as ablation of ⍺vβ6 and ⍺vβ8 in keratinocytes all result in loss of epidermal TRM [21, 30].

We previously reported that local antigen experienced TRM that have encountered cognate antigen in the skin are better able to persist in the epidermis than bystander TRM that have not encountered antigen in the skin when active TGFβ is limited [21]. This is evident when TGFβ activation is experimentally reduced by small molecule inhibition of ⍺vβ6 and ⍺vβ8 and when established TRM compete with newly recruited TEFF cells for limiting amounts of TGFβ activation. In both instances, bystander TRM are preferentially lost from the epidermis while local antigen experienced TRM persist. Thus, encounter with antigen in the skin results in more fit TRM and represents a potential opportunity to preferentially enrich antigen-specific over bystander TRM cells during repeated challenges.

Herein, we delineate mechanisms of TRM homeostasis by demonstrating that within a population of endogenous TRM, antigen encounter in the skin is required for their full differentiation whereas bystander TRM are maintained at an earlier developmental stage. We also find that local antigen experienced TRM have increased proliferative capacity during a recall response compared to bystander TRM that is dependent on affinity of antigen experienced in skin thus providing an additional functional attribute defining TRM fitness. Finally, we show that expression of TGFβRIII by local antigen experienced TRM which can be induced following TCR ligation is required for epidermal persistence when active TGFβ is limiting.

Results

Epidermal TRM are transcriptionally heterogeneous

The recruitment of effector CD8+ T cells into mouse flank skin via a viral infection (e.g. Vaccinia Virus) or through an inflammatory stimulus (e.g. “DNFB-pull”) results in equal numbers of long-lived CD103+ epidermal TRM and is independent of the presence or absence of cognate antigen in the skin [18, 29, 31]. This, however, has only been demonstrated using HSV-specific (gBT-I) or OVA-specific (OT-I) TCR transgenic T cells. To determine if endogenous polyclonal CD8+ T cells share this phenotype, we employed a dual VV infection “DNFB-pull” model. Cohorts of wildtype C57BL/6 mice were infected on the left flank with VV by skin scarification (Figure 1A). The infection expanded CD8+ effectors specific for VV antigens which were then recruited into the left flank in response to the ongoing infection-induced inflammation. DNFB is applied to the right flank 5 days post infection to recruit into the skin circulating CD8 effectors expanded by infection. Mice were then rested for 50+ days to allow for the formation of TRM. We have previously found that evaluation of TRM numbers in the epidermis by flow cytometry is much less accurate than direct immunofluorescent visualization [21].

Epidermal TRM are transcriptionally heterogeneous

(A) Experimental design. Mice were treated with vaccinia virus infection by skin scarification on the left flank on day 0, and then on day 5 post infection the right flank was painted with 0.15% DNFB. On day 56, flanks were harvested for either epidermal whole mount, flow cytometry, or cells were sorted for scRNA-seq. Image created with BioRender.com. (B) Representative images and (C) quantification of epidermal whole mounts of VV and DNFB treated flanks harvested on day 50 post infection stained for CD8a. (D) Representative flow plots gated on CD45+ CD3+ CD8+ CD90.2+ cells isolated from VV or DNFB treated flanks. (E) Representative flow plots and (F) quantification of B8R tetramer binding of TRM isolated from VV or DNFB treated flanks. (G) The integrated Minimal-Distorted Embedding of all 96 experiments from the ImmgenT consortium with annotated clusters. (H) Minimal-Distorted embedding visualization of transcriptional clusters of skin TRM projected over the ImmgenT dataset. (I) The percentage of cells in each transcriptional cluster found in VV or DNFB treated sites. Each symbol represents paired data from the same individual animal (C, F). Data are representative of 3 separate experiments. Scale bar in (B) represents 50um.

Evaluation of whole mounted epidermal sheets stained with anti-CD8 revealed equal numbers of CD8+ T cells at the VV infected (left flank) and DNFB treated (right flank) sites (Figure 1B and C). Expression of CD103 on TRM as evaluated by flow cytometry was also similar at both sites (Figure 1D and S1A). To evaluate the frequency of antigen-specific T cells at each site, skin from a separate cohort was examined by flow cytometry using the B8R tetramer which recognizes an immunodominant epitope of VV in H2-Kb[32]. We observed equivalent numbers of B8R+ TRM at the VV-infected and DNFB treated sites (Figure 1E and F). From these data we conclude that polyclonal CD8 T cells form epidermal TRM with equal efficiency when recruited into the skin by viral infection or sterile inflammation and that the presence or absence of cognate antigen in the skin has no effect on the number or frequency of antigen-specific TRM.

Although TRM efficiently populate the epidermis independently of cutaneous cognate antigen, we have previously demonstrated that TRM that form at skin sites containing cognate antigen are functionally distinct from bystander TRM that form in its absence[21]. To analyze potential transcriptional differences between local antigen experienced and bystander TRM, we performed single-cell RNAseq in collaboration with the ImmgenT consortium [33]. The consortium consists of multiple groups that isolated populations of murine T cells from multiple tissues and contexts and then subjected them to single-cell RNA-seq [33](see Methods). 627,692 mature T cells from 80 different experiments and 703 samples were integrated and batch-corrected using totalVI [34] (see Methods).

Cells were projected in two dimensions using Minimal-Distorted Embedding (see Methods) revealing the expected transcriptional clustering of distinct types of T cells (Figure 1G). As part of the consortium, we isolated cells from the flanks of rested (>65 days) VV-infected or DNFB treated flank skin by enzymatic digestion. Cells were pooled from 10 mice and purified based on expression of CD90 and CD8 (FigureS1B). As expected, most cells expressed CD69 and CD103 (Figure S1C). Skin cells from each group and naïve spleen cells were hash tagged and analyzed by single-cell RNA-seq.

Analysis of transcripts revealed a total of 6 clusters (0-5) which are shown overlayed on the complete ImmgenT cell embedding (Figure 1H, S1D). As expected, most cells clustered with CD8⍺β T cells (Figure 1G, black). There was, however, a high degree of heterogeneity and some, possibly contaminating cells fell outside of this region. A comparison of the relative frequency of cells in each of the clusters (0-5) isolated from the VV or DNFB sites revealed a strong enrichment of cluster 3 in the VV group and of clusters 0 and 2 in the DNFB group (Figure 1I, magenta). This suggested that cluster 3 cells were manifesting a distinct transcriptional state likely induced by antigen encounter in the skin.

Cutaneous antigen is required for complete TRM differentiation

To test the hypothesis that cells in cluster 3 could be TRM that have encountered antigen in the skin during development, we performed signature score enrichment analysis comparing the transcriptional profile of our cells to published core genes of well-defined CD8 T cell states. Signature scores were calculated based on normalized differential gene expression compared to core genes from TRM (GSE47045), activated T cells (TACT, GSE10239), circulating memory T cells (TMEM, GSE41867), and exhausted T cells (TEX, GS41867) [35–38]. Notably, cells in cluster 3 showed enriched expression of TRM core genes compared to all other clusters (Figure 2A). We performed a similar analysis using pseudo-bulk analysis of all cells isolated from the VV or DNFB sites. Consistent with the enrichment of cluster 3 in cells from the VV site, we found a strong enrichment of TRM core genes in the VV group (Figure 2B). These data suggest that cells in cluster 3 isolated from the VV site represent fully differentiated TRM.

Cutaneous antigen is required for complete TRM differentiation

(A) Signature score analysis of skin T cell clusters calculating the enrichment for previously published T cell state gene sets (N=naïve). Signature score = Mean (average upgenes z-scores) - Mean(average downgenes z-scores). *** p < 0.001 by Dunnett’s multiple comparisons of cluster 3 to all other clusters. (B) Pseudobulk analysis comparing signature scores of cells isolated from VV and DNFB sites as in (A). ** p < 0.01 and *** p < 0.001 by paired Student’s t-tests. (C) Signature score analysis of individual clusters or (D) cells isolated from VV and DNFB sites calculating the enrichment for compared to epidermal TRM isolated at the indicated day post infection. ***p < 0.001 by Dunnett’s multiple comparison test of cluster 3 to all other clusters in (C). * p < 0.05, ** p < 0.01, **** p < 0.0001 by paired Student’s t-tests in (D). Datapoints and error bars represent mean and 95% confidence interval of the relative enrichment of each set of DEGs.

To further test the hypothesis that cluster 3 represents fully differentiated TRM, we performed signature score enrichment analysis of clusters 0-5 against a dataset of skin T cells isolated at different times following skin VV infection (GSE79805) [39]. Cells from clusters 0, 2 and 3 showed similar signature scores in comparison with skin T cells up to day 5 post infection, however clusters 0 and 2 then plateaued suggesting a lack of continued differentiation (Figure 2C). In contrast, cluster 3 transcripts scored higher at later time points in TRM differentiation. Signature scores for cells isolated from the DNFB and VV sites were similar at early time points but diverged at later time points with increased scores observed in cells from the VV site (Figure 2D). Taken together, these data suggest that TRM isolated from skin sites where they can encounter cognate antigen are transcriptionally more fully differentiated TRM while bystander TRM isolated from a site lacking cognate antigen remain in a less differentiated state.

Local antigen experienced TRM have improved expansion in a recall response

Analysis of the 10 highest differentially expressed genes (DEGs) in cluster 3 (Figure S1D) showed increased expression of genes associated with T cell activation including Dusp1 and Nr4a1 as well as the AP-1 family members Junb and Fos [40–44]. Given the importance of AP-1 family members in T cell proliferation and differentiation, we next tested whether this corresponded to functional differences in TRM activity [43, 45–48].

An important functional property of TRM is their capacity to rapidly expand following re-encounter with cognate antigen [49, 50]. To determine whether local antigen experienced and bystander TRM have different recall responses, we repeated our dual VV infection and DNFB “pull” model in combination with an OT-I adoptive transfer model to allow for antigen recall using the SIINFEKL peptide. CD90.1+ OT-I cells were adoptively transferred into naïve C57BL/6 mice followed by infection with a vaccinia virus expressing SIINFEKL peptide, OVA257-264 (VV-OVA) on the left flank. On day 5 post infection, the right flank was treated with DNFB to recruit expanded OT-I effectors to the site (Figure 3A). On day 50+ after infection, topical SIINFEKL peptide or control PBS was painted on both flanks in a single application (1° recall). The total number of OT-I cells in the epidermis 6 days later was determined by immunofluorescent microscopic evaluation of the CD90.1 congenic marker in epidermal whole mounts (Figure 3B-C). Restimulation with peptide led to an expansion of TRM at both sites compared to PBS treated controls. Notably, local antigen experienced OT-I TRM cells in the VV-OVA treated flanks expanded to a larger extent compared to bystander TRM in DNFB treated flanks. To inhibit potential recruitment of circulating effector OT-I cells, we repeated these experiments administering FTY720 to block migration of T cells from lymph node or titrated anti-Thy1.1 antibody to ablate circulating OT-I but not TRM (Figure 3D). Both methods successfully depleted OT-I from the blood (Figure S2A-D), without affecting TRM in the skin (Figure S2E). Administration of FTY720 (Figure 3E-F) or anti-Thy1.1 treatment (Figure 3G) had no effect on the number of OT-I TRM in the epidermis after a 1° recall response and local antigen experienced TRM from the VV-OVA treated flank still expanded to a greater degree than bystander TRM from the DNFB treated flank. Thus, the increased number of epidermal OT-I at the local antigen experienced VV site during a 1° recall does not appear to result from recruitment of circulating cells.

Local antigen experienced TRM have improved expansion in a recall response

(A) Experimental design. Mice were adoptively transferred with Thy1.1+ OT-I T cells on day -1, then infected with OVA-expressing vaccinia virus by scarification on the left flank on day 0. On day 5 post-infection the right flank was painted with 0.15% DNFB. On day 50, a primary recall response with SIINFEKL peptide in acetone and olive oil was painted on both flanks and harvested on day 56. In some cohorts, mice were allowed to rest for an additional 120 days and then treated with topical SIINFEKL peptide again on day 176 and harvested on day 182. (B) Representative images and (C) quantification of epidermal whole mounts isolated from VV-OVA and DNFB treated flanks stained for anti-Thy1.1 (n=10 animals). (D) Experimental design. Mice were treated as in (A), but for 6 days prior to SIINFEKL treatment they were given either i.p. FTY720, i.p. PBS, i.v. titrated anti-Thy1.1 or i.v. isotype control. (E) Representative epidermal whole mounts images and (F) quantification of skin from VV-OVA treated flanks on day 6 of a primary recall response treated with either FTY720 or PBS. (G) Quantification of total Thy1.1 OT-I cells in epidermal whole mounts on day 6 after a primary recall response in mice treated with either isotype or anti-Thy1.1 depleting antibody. Each symbol represents paired data from an individual same animal (F, G). Data are representative of 3 independent experiments. *p<0.05 by paired Student’s t-tests. Scale bar represents 50um (B, F). Each symbol represents the mean +/- SEM (C). Panels A and D created with BioRender.com.

Finally, to determine whether increased expansion by local antigen experienced TRM persists with multiple rounds of stimulation, we repeated this experiment but allowed mice to rest for 120 days after the 1° recall response. The number of TRM at both the VV-OVA and DNFB sites contracted down to equivalent numbers that were increased compared to PBS treated, control mice (Figure 3C). Mice were then given a 2° recall by a second application of peptide at day +176. The number of TRM expanded but TRM at the VV-OVA site increased by a larger amount than TRM at the DNFB site (Figure 3B-C). Based on these data, we conclude that TRM encountering antigen in the skin during development expand more efficiently in response to antigen re-encounter and this phenotype persists long-term with subsequent antigen encounters.

Local antigen experienced TRM exhibit increased proliferation during a recall response

We next hypothesized that the increased expansion of local antigen experienced TRM during a recall response resulted from increased proliferation. To test this, as above, WT mice adoptively transferred with OT-I cells were infected with VV-OVA on the left flank and DNFB on the right flank 5 days later. After at least 50 days of rest, both sides were painted with topical SIINFEKL peptide and mice were administered 2 mg of BrdU i.p. daily (Figure 4A). Epidermal sheets for immunofluorescent visualization were harvested 2 days later. We noted a clear increase in the percent and total number of doublets as well as the total number of OT-I T cells in the epidermis at the VV-OVA site compared with the DNFB site (Figure 4B-D). Similar results were obtained by flow cytometry, with an increase in the percentage of BrdU+ in TRM isolated from the VV-OVA site (Figure 4E-G). Staining for Annexin V was equivalent in both groups suggesting that an altered rate of apoptosis does not contribute to the observed changes in expansion (Figure 4H-I). In sum, local antigen experienced TRM show augmented proliferation during an antigen recall response compared with bystander TRM. Moreover, an enhanced proliferative capacity following antigen restimulation can now be added to other parameters of TRM fitness including enhanced epidermal persistence when active TGFβ is limited.

Local antigen experienced TRM have increased proliferation during a recall response

(A) Experimental scheme. Image created with BioRender.com. (B) Representative images of epidermal whole mounts of VV-OVA or DNFB treated flanks on day 2 of a recall response. Arrows highlight cell doublets. (C) Quantification showing percent OT-I cells that are doublets in epidermal whole mounts on day 2 after primary recall response, and (D) total number of OT-I doublets. (E) Representative flow cytometric plots and (F) quantification showing BrdU expression in gated CD45+ CD8+ CD90.1+ CD103+ CD69+ cells isolates from VV-OVA or DNFB treated flanks. (G) Quantification of total numbers BrdU+ OT-I cells combining OT-I numbers with percentage of BrdU incorporation in (F). (H) Representative histograms and (I) quantification of Annexin V expression in OT-I cells isolated from VV-OVA or DNFB treated flanks or OT-I cells heat-treated at 60°C for 1 hour (HK). Data are representative of 3 separate experiments. *p<0.05, **p<0.01 by Students paired t-tests (C), (D), (F) and (G) or Dunnett’s test (I). Scale bar represents 50um.

TRM fitness depends on TCR signal-strength

We next hypothesized that the strength of TCR signal provided in the skin during TRM differentiation should correlate with TRM phenotype at late time points. To test this hypothesis, WT mice were adoptively transferred with OT-I cells then infected with VV-OVA on the left flank and 5 days later were challenged with DNFB on the right flank (Figure 5A). One day later, the DNFB site was painted with PBS, SIINFEKL (N4), or an altered peptide ligand SIYNFEKL (Y3) which has ¼ the avidity for the OT-I TCR receptor [51]. After 50+ days of rest, all mice were challenged in a 1° recall response with topical N4 or control PBS at the DNFB sides (Figure 5A). Epidermal sheets for immunofluorescence were harvested 6 days later. As expected, mice restimulated with PBS showed equivalent numbers of TRM regardless of which peptide was provided during development (Figure 5B-C). Following a 1° recall response, TRM exposed to N4 (DNFB + N4) expanded to a level equivalent to TRM from the VV-OVA site. In contrast, TRM exposed to Y3 (DNFB + Y3) showed a degree of expansion that was intermediate compared to TRM that did not experience antigen in the skin (DNFB+PBS) and those that experienced SIINFEKL (DNFB + N4). Thus, the strength of TCR engagement determines the fitness of TRM based on their capacity to expand during a 1° recall response.

TRM fitness depends on TCR signal-strength

(A) Experimental scheme. Image created with BioRender.com. (B) Representative images and (C) quantification of epidermal whole mounts (CD90.1 cyan) of VV-OVA and DNFB+ APL treated flanks, at steady state (no recall) or 6 days post recall response. Each symbol represents data from an individual animal. Data are representative of 5 independent experiments. *p<0.05 by unpaired Student’s t-tests. Scale bar, 50um.

Local antigen experienced TRM have improved persistence mediated by TGFßRIII

A feature of local antigen experienced TRM is that they are better able to persist in the epidermis than bystanders when levels of activated TGFβ are limited, either artificially or during competition with newly recruited TEFF [21]. We have previously found that expression of the canonical TGFβ receptors, Tgfbr1 and Tgfbr2 have equivalent expression in local antigen experienced and bystander TRM [21]. However, there is a third TGFβ receptor, TGFβRIII, that lacks a signaling component but functions as a reservoir of ligand for TGFβ, increasing avidity of the TGFβ receptor [52, 53]. Notably, expression of TGFβRIII has been reported to increase in T cells following TCR ligation [54]. This was confirmed in vitro with anti-CD3 anti-CD28 stimulated OT-I splenocytes (Figure 6A). We also observed increased expression of TGFβRIII in vivo comparing local antigen experienced OT-I TRM with bystanders (Figure 6B-C). Based on this data, we hypothesized that increased expression of TGFβRIII on antigen experienced TRM could explain their capacity to maintain epidermal residence when available TGFβ is limiting.

Local antigen experienced TRM have improved persistence mediated by TGFßRIII

(A) Representative flow cytometric plots of TGFßRIII staining of Thy1.1+ OT-I cells stimulated in vitro for 48 hours with PBS or anti-CD3 anti-CD28. (B) Representative histograms showing TGFßRIII expressing in CD45+ CD3+ CD8+ CD90.1+ gated OT-I cells isolated from VV-OVA, DNFB or untreated flanks at least 50 days post infection. (C) Quantification of (B). (D) Schematic demonstrating genetics of Tgfbr3WT and Tgfbr3ΔCD8 mice. (E) Representative histogram and (F) quantification of TGFßRIII expression in huNGFR- or huNGFR+ Tgfbr3ΔCD8 T cells harvested 5 days following i.p. treatment with tamoxifen then stimulated in vitro for 48 hours with anti-CD3 and anti-CD28. (G) Experimental scheme. (H) Representative histogram of CD45.2+CD3+CD8+CD90.1+ OT-I cells isolated from LNs after tamoxifen treatment demonstrating transformation efficiency. (I) Representative epidermal whole mounts showing Thy1.1 staining (green), huNGFR staining (red) or merge (yellow) of VV-OVA or DNFB treated flanks from mice adoptively transferred with either Tgfbr3WT or Tgfbr3ΔCD8 cells, treated with tamoxifen i.p., and then given either PBS or i.p. CWHM12 for 10 days. Hair follicles in the sample present as long yellow streaks. (J) Quantification of total huNGFR+ Thy1.1+ OT-I in (I). Each symbol represents data from an individual animal. Data is representative of 3 independent experiments. *p<0.05 by Dunnett’s test (C) or paired Student’s t-tests (F) and (J). Scale bar represents 50um. Panels D and G created with BioRender.com.

To test this, we generated OT-I Thy1.1+ E8i-creERT2 huNGFR Tgfbr3fl/fl (Tgfbr3ΔCD8) mice which allow for inducible ablation of Tgfbr3 from TRM as well control OT-I Thy1.1+ E8i-creERT2 huNGFR Tgfbr3WT/WT (Tgfbr3WT) mice (Figure 6D). To validate these mice, Tgfbr3ΔCD8 mice were treated with 0.05mg/g tamoxifen i.p. for 5 days. CD8+ T cells from tamoxifen treated Tgfbr3ΔCD8 mice were then isolated from spleen and lymph nodes and cultured for 2 days with anti-CD3 anti-CD28. Cells were evaluated by flow cytometry for expression TGFβRIII and huNGFR, an indicator of successful induction of cre-mediated excision by tamoxifen. In the huNGFR-negative population where cre was not activated, approximately 50% of the cells expressed TGFβRIII (Figure 6E-F). In contrast, TGFβRIII was largely absent from cells co-expressing huNGFR indicative of efficient ablation of Tgfbr3.

To ablate TGFβRIII from TRM once they have fully differentiated in the epidermis, we next adoptively transferred CD8+ T cells isolated from either Tgfbr3WT or Tgfbr3ΔCD8 mice into naïve WT mice followed by skin VV-OVA infection on the left flank. On day 5 post infection, the right flank was treated with DNFB. After at least 50 days of rest to allow for TRM formation, both cohorts were treated daily with 0.05mg/g tamoxifen i.p. to transform approximately 50% of adoptively transferred cells, generating populations of both control huNGFR negative and huNGFR positive OT-I cells. To test the effects of TGFβRIII ablation, TGFβ activation was inhibited for 10 days by i.p. administration of CWHM12, a small molecule inhibitor of integrins ⍺vβ6 and ⍺vβ8 (Figure 6G-H).

Analysis of epidermal whole mounts by immunofluorescence revealed that blockade of TGFβ activation reduced the number of bystander Tgfbr3WT TRM at the DNFB site, but not local antigen experienced TRM at the VV-OVA site, consistent with our earlier results (Figure 6I-J). In contrast, local antigen experienced huNGFR+ Tgfbr3ΔCD8 TRM from the VV-OVA site that lack TGFβRIII were efficiently depleted from the epidermis by CWHM12 treatment. The depletion of bystander huNGFR+ Tgfbr3ΔCD8 TRM at the DNFB site was augmented. Notably, numbers of Tgfbr3ΔCD8 TRM in cohorts treated with vehicle remained at normal levels indicating that loss of TGFβRIII does not affect TRM epidermal residence in the steady state. Thus, expression of TGFβRIII by local antigen experienced TRM which is downstream of TCR ligation is required for their capacity to remain in the epidermis when activated TGFβ is limiting.

Discussion

Herein we have demonstrated that the increased capacity of local antigen experienced TRM to persist in the epidermis when levels of TGFβ are limited is mediated by increased expression of TGFβRIII. We also show that local antigen experienced TRM have increased proliferative capacity during repeated antigen recalls. In addition, the increased proliferative capacity was directly correlated with the strength of TCR stimulation during TRM development. Finally, we found that local antigen experienced TRM appear more transcriptionally related to fully differentiated TRM. Taken together, these data support a model in which TCR engagement by cognate antigen in the skin is a required final step in TRM differentiation resulting in their increased fitness exemplified by increased proliferative capacity and the ability to persist in the epidermis when active TGFβ is limited.

We propose that the augmented fitness of local antigen experienced TRM represents a mechanism to enrich for high avidity TCR-clones in the epidermis. Skin inflammation recruits TEFF into the skin, some of which develop into TRM. In the absence of competition with pre-existing TRM, TRM form equally efficiently in the presence or absence of cognate antigen. Thus, we found equivalent numbers of bystander TRM at the DNFB and local antigen experienced TRM at the VV sites with both TCR transgenic and endogenous T cells. In contrast, when new TEFF are recruited into sites with pre-existing TRM, there is clonal competition for limited amounts of active TGFβ resulting in enrichment of fitter, local antigen experienced TRM [21]. We now find that this enrichment likely results from 2 different competitive advantages. First, antigen encounter in the skin results in increased expression of TGFβRIII which increases TGFβ avidity for the signaling by the TGFβ receptor. Since TGFβ signaling is required for epidermal persistence, this would provide an advantage for local antigen experienced TRM over bystanders. Second, local antigen experienced TRM have increased proliferation when re-encountering antigen in the epidermis. Following repeated challenges which would be expected outside of SPF conditions, the combination of improved expansion and persistence would work together to enrich for high avidity clones thereby shaping the epidermal CD8+ T cell memory pool. Recently, it has been observed that TRM can contribute significantly to the pool of circulating memory cells [55–59]. Thus, mechanisms augmenting epidermal TRM fitness that shape the pool of epidermal TRM may also affect the pool of systemic memory cells and represent an example of extra-thymic clonal section.

When TRM were challenged in a primary antigen recall response, we noted that local antigen experienced TRM expanded to a greater extent than bystanders. This expansion resulted from increased in-situ proliferation with minimal contribution from newly recruited TEFF, consistent with prior reports [20, 60, 61]. Interestingly, a similar phenomenon occurred following a second encounter with antigen. This indicates that encounter with peptide at a late time point after TRM differentiation (>50 days) is insufficient to convert bystander TRM into local antigen experienced TRM. Thus, there appears to be a window during TRM development when TCR engagement can allow for full differentiation. We also observed after a single recall response that TRM contracted to an elevated baseline suggesting an increase in the epidermal niche. We speculate this may result from a reduced T cell intrinsic requirement for survival and/or homeostatic proliferation factors such as IL-7 or IL-15 or increased expression of these factors by keratinocytes[62, 63]. Altered sensitivity or availability of TGFβ is unlikely to explain the increased niche size as this would be predicted to vary between local antigen experienced and bystander.

Transcriptional analysis of TRM isolated from the small intestine have revealed intra-organ heterogeneity, with unique transcriptional populations arising early during TRM development [64–66]. This aligns well with our identification of 6 distinct transcriptional clusters of epidermal TRM. Cluster 3 appears to represent fully differentiated TRM based on comparison with other TRM datasets. In addition, cluster 3 cells more highly expressed the activation and proliferation associated genes Junb, Fos and Dusp1 as well as Nr4a1. Increased basal expression of the AP-1 family members Junb and Fos could contribute to the enhanced proliferation of antigen-experienced epidermal TRM during a recall response. Intriguingly, memory CD8 T cells lacking the transcription factor Zbtb20 manifest elevated expression of AP-1 family members and mount more robust antitumor responses [71]. The Nr4a1 gene encodes for Nur77, which is induced by TCR signaling and its expression correlates with peptide avidity. Notably Nur77 is required for TRM formation in the liver [23–25, 67, 68]. Interestingly, cells in cluster 3 only accounted for 27% of TRM that had the opportunity to encounter their cognate antigen in the VV treated flank. We speculate that not all clones at the VV site fully develop into fitter TRM due to lower TCR avidity or specificity to viral antigens only expressed early during infection which would be absent once the clones arrived into skin.

In sum, TCR signaling during TRM differentiation represents a previously unappreciated final step in TRM differentiation. This results in fitter TRM with a lower requirement for TGFβ transactivation due to increased expression of TGFβRIII and enhanced proliferation in response to peptide stimulation. Moreover, the differing responses to altered peptide ligands indicate that the degree of fitness depends on TCR signal-strength. Thus, polyclonal TRM likely develop into a spectrum of bystander to local antigen experienced cells based on TCR avidity. Though we have focused entirely on epidermal T cells, we suspect that these mechanisms may play a role in other epithelial tissues where residency is also dependent upon TGFβ. Additionally, we have solely investigated memory CD8+ T cells after acute inflammation, the role of ongoing TCR-engagement during chronic antigen encounter remains unexplored.

Materials and Methods

Mice

We generated Thy1.1+Rag−/−OT-I mice by crossing OT-I mice with Rag−/− and Thy1.1 mice. E8I-creERT2 and ROSA26.LSL.hNGFR reporter mice were developed by Dario A.A. Vignali (University of Pittsburgh) [29]. E8I-creERT2 mice and Thy1.1+Rag-/- OT-I mice were crossed with ROSA26.LSL.hNGFR reporter mice to generate Tgfbr3WT mice and additionally with Tgfbr3loxP mice to obtain Tgfbr3ΔCD8 mice. Tgfbr3loxp mice were developed by Herbert Y Lin (Program in Membrane Biology/Nephrology, Department of Medicine, Massachusetts General Hospital and Harvard Medical School, Boston, Massachusetts) [69]. We used age- and sex-matched female mice that were between 6 and 12 weeks of age at the start of all experiments. All mice were maintained under specific-pathogen-free conditions and all animal experiments were approved by University of Pittsburgh Institutional Animal Care and Use Committee.

TRM cell models and blocking TGFβ activation treatments

Mice were infected by skin scarification (skin infection) with 3 × 106 plaque-forming units recombinant vaccinia virus expressing the SIINFEKL peptide of ovalbumin (VV-OVA), or vaccinia virus without recombinant peptide expression (VV). For skin scarification, 45 μL of VV-OVA or VV was applied to shaved left flank (4–5 cm2) and the skin were gently scratched 100 times with 27 G needle under anesthesia. VV or VV-OVA infected mice were further treated with 0.15% 1-Fluoro-2,4-dinitrobenzene (DNFB, D1529: Sigma-Aldrich) in 4:1 acetone : olive oil (O1514, Sigma-Aldrich) on the flank opposite the site of infection (40 μL) 5 days post infection. In some experiments, 4:1 acetone and olive oil was added to the stock solution of OVA257–264 (SIINFEKL or SIYNFEKL, RP101611, Genscript/Fisher) in DMSO (D2650, Sigma-Aldrich) (10 mg/mL) for a final concentration of OVA257–264 at 0.2 mg/mL and 50 μL was painted to the DNFB-treated skin 1 day after DNFB treatment.

Recall response

50+ days after VV-OVA infection. 4:1 acetone and olive oil was added to the stock solution of OVA257–264 (SIINFEKL, RP101611, Genscript/Fisher) in DMSO (D2650, Sigma-Aldrich) (10 mg/mL) for a final concentration of OVA257–264 at 0.2 mg/mL and 50 μL was painted to the both flanks.

CWHM12 treatments

For blocking TGFβ activation with a small molecule integrin inhibitor, compound CWHM12 (kindly provided by Indalo therapeutics, Cambridge, MA) was solubilized in 100% DMSO. Further dilution to 50% DMSO was made in sterile 100% PBS and dosed to 100 mg per kg body weight per day delivered by i.p. injections once a day for 10 days.

Adoptive transfers

OT-I cells were purified from spleen and lymph nodes by MojoSort Mouse CD8 T Cell Isolation Kit (48007, Biolegend) according to the manufacturer’s instructions and 1 × 105 OT-I cells were intravenously transferred is all experiments. All mice were allowed to rest for 1 day before further experimentation.

Tamoxifen treatment

Tamoxifen (T5648; Sigma-Aldrich) was dissolved in 1/10th volume of 200 proof ethanol following incubation at 37 °C for 15–30 min with 300 rpm shaking. Corn oil (C8267, Sigma-Aldrich) was added for a final concentration of Tamoxifen at 10 mg/ml and was administered to mice for 5 consecutive days by intraperitoneal injection at 0.05 mg per g body weight, in order to transform approximately 50% of adoptively transferred cells.

BrdU treatment

BrdU (B5002, Sigma) was dissolved in sterile PBS. 2 mg in 200uL of PBS was injected i.p. daily into each mouse for 2 consecutive days before harvest.

Immunofluorescence of epidermis

Epidermal sheets were prepared as previously described [30]. Briefly, epidermal side of shaved defatted flank skin was affixed to slides with double-sided adhesive (3M, St. Paul, MN). Slides were incubated in 10 mM EDTA in PBS for 90 min at 37°C, followed by physical removal of the dermis. The epidermal sheets were fixed in 4% PFA at RT for 15 min. The epidermal sheets were blocked with PBS containing 0.1% tween-20, 2% BSA and 2% rat serum for 1 hour at RT before staining 1 hour with antibodies at RT in PBS containing 0.1% tween-20 and 0.5% BSA. The slides were mounted with ProLong™ Gold Antifade Mount with DNA stains DAPI (P36931, Thermofisher). Images were captured on Keyence BZX800 fluorescent microscope (Keyence, Osaka, Japan). Analysis was performed using BZ-H4A Advanced Analysis Software(Keyence, Osaka, Japan). For enumeration of cells, three images from distant sites within an epidermal sheet from a mouse were counted (total 3 mm2) manually after image processing by Adobe Photoshop (version 6) and the average number per mm2 epidermis was calculated as representative of the epidermal sheet. Anti-CD8α (53–6.7), Thy1.1 (OX-7) and huNGFR (ME20.4), were purchased from Biolegend.

Depletion of circulation cells

For depletion of circulating memory OT-I cells, OT-I adoptive transfer VV-OVA and DNFB treated mice were injected i.p. with 0.3–1 μg anti-Thy1.1 (HIS51, ThermoFisher) in 200 μL PBS for 10 consecutive days. Depletion of OT-I T cells (< 0.5% of total CD8+ T cells and < 1%) were confirmed by staining with anti-Thy1.1 (OX-7) using blood 3 days after depletion. For FTY720 treatment, OT-I adoptive transfer VV-OVA and DNFB treated mice were injected i.p. with FTY720 (10006292, Fisher/Cayan Chemical) at 1 μg/g body weight in sterile PBS with 0.5% DMSO (D2650, Sigma-Aldrich) daily for 10 days. Depletion of circulating cells was confirmed with staining of anti-Thy1.1 (OX-7) and anti-CD8α (53–6.7) using blood 3 days after the end of treatment.

Flow cytometry

Preparation of single cells suspension from skin, shaved skin was harvested, and fat tissues was removed by forceps mechanically. The skin was then mechanically digested with either scissors or a gentleMACS Dissociator (Miltenyi Biotec, Bergisch Gladbach, Germany), and then resuspended in RPMI 1640 media (Gibco, Grand Island, NY) containing 2.5 mg/ml Collagenase XI (9001-12-1, Sigma-Aldrich), 0.1 mg/ml DNase (04536282001, Sigma-Aldrich), 0.01 M 4-(2-hydroxyethyl)-1-piperazineethanesulfonic acid (7365-45-9, Sigma-Aldrich), and 10% fetal bovine serum, followed by incubation in a shaking incubator for 30 minutes at 37 °C. The resulting cell mesh was filtered through a 40-μm cell strainer (BD Biosciences, San Jose, CA). The single-cell suspension was then resuspended in 44% percoll (89428-524, VWR) in RPMI and layered over 67% percoll in PBS. After 20 minutes of centrifugation at 931 x g at RT, the interface was isolated and resuspended in FACs buffer. LNs (axillary and inguinal) were incubated in 400 U/ml Collagenase D (Roche Applied Science, Penzberg, Germany) and 0.1 mg/ml DNase in RPMI 1640 with 10% fetal bovine serum for 40 minutes at 37 °C and then minced through a 40-μm cell strainer. In some experiments LNs were heat-treated at 60°C in a hot water bath for 1 hour. Blood was collected with heparin (H3393, Sigma-Aldrich) and treated with red blood cell lysing buffer (R7757, Sigma-Aldrich). Single-cell suspensions were blocked with 2.4G2 culture supernatant (ATCC, Manassas, VA). Surface staining was performed in standard FACS buffer for 30 minutes at 4 °C. Anti-CD8α (53–6.7), CD44 (IM7), CD3 (17A2), Thy1.2 (30-H12), Thy1.1 (OX-7), CD69 (H1.2F3), CD103 (2E7), BrdU (3D4), MHC-II (M5/114.15.2), CD45.2 (104) and Annexin V were purchased from Biolegend. Anti-TGFßRIII (1C5H11) was purchased from Sigma-Aldrich. Soluble tetrameric B8R20–27/H2-Kb complex (made by the NIH Tetramer Core Facility) was conjugated to PE-labeled streptavidin. A BD LSRFORTESSA (BD Biosciences) and Flowjo software (TreeStar, Ashland, OR) were used for analysis.

Single-cell RNA and Totalseq C-Sequencing - Sample Preparation

Flow cytometry and HT antibody staining

Single-cell lymphocyte suspensions were isolated as described above and stained with anti-CD45-APC, anti-CD8a, anti-CD44, anti-CD103 and different TotalSeq-C Anti-Mouse Hashtags (TotalSeq-C Anti-Mouse Hashtags 1-10 Biolegend #155861-155879, in staining buffer (DMEM, 2% FCS and 10mM HEPES)) for 20min at 4°C in the dark. Standard spleen cell suspensions were stained under identical conditions but included anti-CD45-FITC (BioLegend 109806, clone 104) to differentiate them from other samples when pooled. Each sample was labeled with a unique hashtag, enabling the downstream assignment of individual cell (10x cell barcode) to their respective source samples [70].

Cell sorting

For each sample, 500-5000K DAPI- CD45+ Thy1.2+ CD8a+ (skin samples) or DAPI- CD45+ Thy1.2+ CD8a+ CD44+ CD103+ (LN samples) T cells were sorted on the FACSAria III and pooled in a single collection tube at 4C. DAPI was added just before the sort.

ImmGenT TotalSeq-C custom mouse panel staining

The pooled single cell suspension was stained with the ImmGenT TotalSeq-C custom mouse panel, containing 128 antibodies (Biolegend Part no. 900004815) and FcBlock (Bio X Cell #BE0307). Because 500,000 cells are required for staining, unstained splenocytes were spiked in to reach a total of 500,000 cells.

Second sort

Cells were subsequently sorted a second time with the addition of DAPI to select for live T cells, and include 5,000 total splenocyte standard cells were sorted into a single collection tube.

Single-cell RNA and TotalseqC-Sequencing - Library Preparation.

Cell encapsulation and cDNA library

Single-cell RNA sequencing was performed using the 10x Genomics 5’ v2 platform with Feature Barcoding for Cell Surface Protein and Immune Receptor Mapping, adhering to the manufacturer’s guidelines (CG000330). After cell encapsulation with the Chromium Controller, reverse transcription and PCR amplification were performed in the emulsion. From the amplified cDNA library, smaller fragments containing TotalSeq-C-derived cDNA were separated for Feature Barcode library construction, while larger fragments containing transcript-derived cDNA were preserved for TCR and Gene Expression library generation. Library sizes for both cDNA fractions were evaluated using the Agilent Bioanalyzer 2100 High Sensitivity DNA assay and quantified with a Qubit dsDNA HS Assay kit on a Qubit 4.0 Fluorometer.

RNA library construction

After enzymatic fragmentation and size selection of the cDNA, the library was ligated to an Illumina R2 sequence and indexed using unique Dual Index TT set A index sequences, SI-TT-B6.

TotalSeq-C library construction

Totalseq-C derived cDNA was processed into the Feature Barcode libraries following the manufacturer’s protocol. The library was indexed with a unique dual index TN set TN set A (10x part no. 3000510) index sequence SI-TN-F9.

Sequencing

The three libraries were pooled based on molarity in the following proportions: 47.5% RNA, 47.5% Feature Barcode, and 5% TCR. The pooled libraries were sequenced on an Illumina NovaSeq S2 platform (100 cycles) using the 10x Genomics specifications: 26 cycles for Read 1, 10 cycles for Index 1, 10 cycles for Index 2, and 90 cycles for Read 2.

Single-cell RNA, TCR and TotalseqC-sequencing - Data processing

Code

Code is available on https://github.com/immgen/immgen_t_git/

Count matrices

Gene and TotalSeq-C antibody (surface protein panel and hashtags) counts were obtained by aligning reads to the mm10 (GRCm38) mouse genome using the M25 (GRCm38.p6) Gencode annotation and the DNA barcodes for the TotalSeq-C panel. Alignment was performed with CellRanger software (v7.1.0, 10x Genomics) using default parameters. Cells were identified and separated from droplets with high RNA and TotalSeq-C counts by determining inflection points on the total count curve, using the barcodeRanks function from the DropletUtils package.

Sample demultiplexing

Sample demultiplexing was performed using hashtag counts and the HTODemux function from the Seurat package (Seurat v4.1). Doublets (droplets containing two hashtags) were excluded, and cells were assigned to the hashtag with the highest signal, provided it had at least 10 counts and was more than double the signal of the second most abundant hashtag. Hashtag count data were also visualized using t-SNE to ensure clear separation of clusters corresponding to each hashtag. All single cells from the gene count matrix were uniquely matched to a single hashtag, thereby linking them unambiguously to their original sample.

Quality control (QC) and batch correction

Cells meeting any of the following criteria were excluded from the analysis: fewer than 500 RNA counts, dead cells with over 10% of counts mapping to mitochondrial genes, fewer than 500 TotalSeq-C counts, or positivity for two isotype controls (indicating non-specific TotalSeq-C antibody binding). Non-T cells were excluded based on the expression of the MNP gene signature, B cell signature, ILC gene signature, and the absence of T cell gene signature (score calculated using AddModuleScore_UCell). CITEseq data did not meet quality control and was not used in the analysis.

ImmGen T integration

The data was integrated with the rest of the ImmgenT dataset using the SCVI.TOTALVI model (v1.2.0) and the 10x lane as a batch parameter. Dimensionality reduction was performed using the pymde.preserve_neighbors() function with default parameters [CITATION: https://pymde.org/citing/index.html]. Cell clustering was carried out using the FindClusters() function in Seurat. Manual annotation by the immgenT consortium was done using protein and RNA expression of Cd3e, Trbc1, Trbc2, Cd4, Cd8a, Cd8b1, Foxp3, Mki67, Sell, Cd44, Trgc1, Itgax, Itgam, Ms4a1, CD3, TCRB, THY1.2, CD4, CD8A, CD8B, CD62L, CD44, TCRGD, TCRVG1.1, TCRVG2, TCRVG3, CD19, CD20, ITAM.CD11B, ITAX.CD11C, KLRBC-NK1.1. Data was visualized using the Rosetta software (https://cbdm.connect.hms.harvard.edu/ImmgenT/PublicRosetta/). ImmgenT integration and annotation available on https://www.immgen.org/ImmGenT/.

Clustering and dimensionality reduction

Using Seurat v4.2. [https://pubmed.ncbi.nlm.nih.gov/31178118/], the variance-stabilizing transformation (VST method) was applied, and PCA was conducted on the top 2000 genes. Principal components explaining 80% of the total variance were selected for two-dimensional reduction using UMAP. Clustering was performed on these principal components using the FindClusters() function in Seurat.

Differential Expression

To determine differentially expressed genes, FindMarkers() within Seurat was used. AddModuleScore() was used to visualize aggregated expression of a set of genes. Volcano plots were created using EnhancedVolcano (v1.14.0).

Signature score analysis

Signature score was calculated by first generating differentially expressed genes of each cluster or condition compared to naïve T cells from the spleen control, and then calculating an average z-score for every up and down gene in a core gene signature datasets. The core gene datasets were generated by Jaiswal et al. [35] by calculating the exclusive expression of genes from T-cells in the datasets GSE47045 [23], GSE10239 [37] and GSE41867 [38]. Comparative programs for TRM development over time were analyzed against the transcripts taken from the top 100 differentially expressed genes of each TRM timepoint/ naïve from GSE79805 [39] by adapting published scoring methods with this data set Jaiswal et al. [35]. The overall signature score was scored based on expression data that was quantified using RSEM in transcripts per million (TPM), then log-transformed as log2(TPM+1). It was then centered for each gene across all cells. The centered data was averaged across sets of genes to define signature scores. We subtracted a control score from the signature score, which is defined using the same process on randomly selected gene-sets. A randomized control was generated by the average enrichment of each group compared to a randomly generated gene set of an equal number of probes, as iterated Jaiswal et al. [35].

Quantification and statistical analysis

Groups were compared with Prism software (GraphPad) using two-tailed paired Student’s t-test for comparison of left and right flanks, two-tailed unpaired Student’s t-test for comparison between animals, or Dunnett’s test for comparisons of more than two groups. Data is presented as each data point and mean with the standard error of the mean (s.e.m.). p < 0.05 was considered significant.

eLife Assessment

Weiss et al. provide important new insights and convincing evidence to further our mechanistic understanding of how antigen presentation shapes skin persistence of CD8+ TRM. Using a mouse model for inducible genetic ablation of transforming growth factor beta receptor 3 (TGFBR3) in CD8+ T cells, they demonstrate TGFBR3's role in regulating CD8+ TRM persistence in skin. Furthermore, they show that the strength of T cell receptor (TCR) engagement upon initial CD8+ TRM skin seeding has a positive influence on subsequent TRM expansion following a secondary antigen-reencounter. Together, these mechanisms add to our understanding of how the skin CD8+ T cell repertoire is dynamically responsive to topical antigen.

Reviewer #1 (Public review):

Summary:

Weiss et. al. seek to delineate the mechanisms by which antigen-specific CD8+ T cells outcompete bystanders in the epidermis when active TGF-b is limiting, resulting in selective retention of these cells and more complete differentiation into the TRM phenotype.

Strengths:

They begin by demonstrating that at tissue sites where cognate antigen was expressed, CD8+ T cells adopt a more mature TRM transcriptome than cells at tissue sites where cognate antigen was never expressed. By integrating their scRNA-Seq data on TRM with the much more comprehensive ImmGenT atlas, the authors provide a very useful resource for future studies in the field. Furthermore, they conclusively show that these "local antigen-experienced" TRM have increased proliferative capacity and that TCR avidity during TRM formation positively correlates with their future fitness. Finally, using an elegant experimental strategy, they establish that TCR signaling in CD8+ T cells in epidermis induces TGFBRIII expression, which likely contributes to endowing them with a competitive advantage over antigen-inexperienced TRM.

Weaknesses:

The main weakness in this paper lies in the authors' reliance on a single model to derive conclusions on the role of local antigen during the acute phase of the response by comparing T cells in model antigen-vaccinia virus (VV-OVA) exposed skin to T cells in contralateral skin exposed to DNFB 5 days after the VV-OVA exposure. In this setting, antigen-independent factors may contribute to the difference in CD8+ T cell number and phenotype at the two sites. For example, it was recently shown that very early memory precursors (formed 2 days after exposure) are more efficient at seeding the epithelial TRM compartment than those recruited to skin at later times (Silva et al, Sci Immunol, 2023). DNFB-treated skin may therefore recruit precursors with reduced TRM potential. In addition, TRM-skewed circulating memory precursors have been identified (Kok et al, JEM, 2020), and perhaps VV-OVA exposed skin more readily recruits this subset compared to DNFB-exposed skin. Therefore, when the DNFB challenge is performed 5 days after vaccinia virus, the DNFB site may already be at a disadvantage in the recruitment of CD8+ T cells that can efficiently form TRM. In addition, CD8+ T cell-extrinsic mechanisms may be at play, such as differences in myeloid cell recruitment and differentiation or local cytokine and chemokine levels in VV-infected and DNFB-treated skin that could account for differences seen in TRM phenotype and function between these two sites. Although the authors do show that providing exogenous peptide antigen at the DNFB-site rescues their phenotype in relation to the VV-OVA site, the potential antigen-independent factors distinguishing these two sites remain unaddressed. In addition, there is a possibility that peptide treatment of DNFB-treated initiates a second phase of priming of new circulatory effectors in the local-draining lymph nodes that are then recruited to form TRM at the DFNB-site, and that the effect does not solely rely on TRM precursors at the DNFB-treated skin site at the time of peptide treatment.

Secondly, although the authors conclusively demonstrate that TGFBRIII is induced by TCR signals and required for conferring increased fitness to local-antigen-experienced CD8+ TRM compared to local antigen-inexperienced cells, this is done in only one experiment, albeit repeated 3 times. The data suggest that antigen encounter during TRM formation induces sustained TGFBRIII expression that persists during the antigen-independent memory phase. It remains unclear why only the antigen encounter in skin, but not already in the draining lymph nodes, induces sustained TGFBRIII expression. Further characterizing the dynamics of TGFBRIII expression on CD8+ T cells during priming in draining lymph nodes and over the course of TRM formation and persistence may shed more light on this question. Probing the role of this mechanism at other sites of TRM formation would also further strengthen their conclusions and enhance the significance of this finding.

Reviewer #2 (Public review):

Summary:

The authors set out to dissect the mechanistic basis of their previously published finding that encountering cutaneous antigen augments the persistence of CD8+ memory T cells that enter skin (TRM) (Hirai et al., 2021, Immunity). Here they use the same murine model to study the fate of CD8+ T cells after antigen-priming in the lymph nodes, (1) those that re-encounter antigen in the skin via vaccinia virus (VV) versus (2) those that do not encounter antigen in skin but rather are recruited via topical dinitrofluorobenzene (DNFB) (so-called "bystander TRM"). The authors' previous publication establishes that this first group of CD8+ TRM has a persistence advantage over bystander TRM under TGFb-limiting conditions. The current paper advances this finding by elucidating the role of TGFBR3 in regulating CD8+ TRM skin persistence upon topical antigen exposure. Key novelty of the work lies in the generation and use of the CD8+ T cell-specific TGFBR3 knockout model, which allows them to demonstrate the role of TGFBR3 in fine-tuning the degree of CD8+ T cell skin persistence and that TGFBR3 expression is promoted by CD8+ TRM encountering their cognate antigen upon initial skin entry. Future work directly measuring active TGFb in the skin under different conditions would help identify physiologic scenarios that yield active TGFb-limiting conditions, thus establishing physiologic relevance.

Strengths:

Technical strengths of the paper include (1) complementary imaging and flow cytometry analyses, (2) integration of their scRNA-seq data with the existing CD8+ TRM literature via pathway analysis, and (3) use of orthogonal models where possible. Using a vaccina virus (VV) model, with and without ovalbumin (OVA), the authors investigate how topical antigen exposure and TCR strength regulate CD8+ TRM skin recruitment and retention. The authors use both FTY720 and a Thy1.1 depleting antibody to demonstrate that skin CD8+ TRM expand locally following both a primary and secondary recall response to topical OVA application.

A conceptual strength of the paper is the authors' observation that TCR signal strength upon initial TRM tissue entry helps regulate the extent of their local re-expansion on subsequent antigen re-exposure. They achieved this by applying peptides of varying affinity for the OT-I TCR on the DNFB-exposed flank in tandem with initial VV-OVA + DNFB treatment. They then measured TRM expansion after OVA peptide rechallenge, revealing that encountering a higher-affinity peptide upon skin entry leads to greater subsequent re-expansion. Additionally, by generating an OT-I Thy1.1+ E8i-creERT2 huNGFR Tgfbr3fl/fl (Tgfbr3∆CD8) mouse, the authors were able to elucidate a unique role for TGFBR3 in CD8+TRM persistence when active TGFb in skin is limited.

Weaknesses:

Overall, the authors' conclusions are well supported, although there are some instances where additional controls, experiments, or clarifications would add rigor. The conclusions regarding skin-localized TCR signaling leading to increased skin CD8+ TRM proliferation in-situ and increased TGFBR3 expression would be strengthened by assessing skin CD8+ TRM proliferation and TGFBR3 expression in models of high versus low avidity topical OVA-peptide exposure. The authors could further increase the novelty of the paper by exploring whether TGFBR3 is regulated at the RNA or protein level. To this end, they could perform analysis of their single-cell RNA sequencing data (Figure 1), comparing Tgfbr3 mRNA in DNFB versus VV-treated skin.

For clarity, when discussing antigen exposure throughout the paper, it would be helpful for the authors to be more precise that they are referring to the antigen in the skin rather than in the draining lymph node. A more explicit summary of some of the lab's previous work focused on CD8+ TRM and the role of TGFb would also help readers better contextualize this work within the existing literature on which it builds.

For rigor, it would be helpful where possible to pair flow cytometry quantification with the existing imaging data. Additional controls, namely enumerating TRM in the opposite, untreated flank skin of VV-only-treated mice and the treated flank skin of DNFB-only treated mice, would help contextualize the results seen in dually-treated mice in Figure 1. In figure legends, we suggest clearly reporting unpaired T tests comparing relevant metrics within VV or DNFB-treated groups (for example, VV-OVA PBS vs VV-OVA FTY720 in Figure 3F). Finally, quantifying right and left skin draining lymph node CD8+ T cell numbers would clarify the skin specificity and cell trafficking dynamics of the authors' model.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Weaknesses:

The main weakness in this paper lies in the authors' reliance on a single model to derive conclusions on the role of local antigen during the acute phase of the response by comparing T cells in model antigen-vaccinia virus (VV-OVA) exposed skin to T cells in contralateral skin exposed to DNFB 5 days after the VV-OVA exposure. In this setting, antigen-independent factors may contribute to the difference in CD8+ T cell number and phenotype at the two sites. For example, it was recently shown that very early memory precursors (formed 2 days after exposure) are more efficient at seeding the epithelial TRM compartment than those recruited to skin at later times (Silva et al, Sci Immunol, 2023). DNFB-treated skin may therefore recruit precursors with reduced TRM potential. In addition, TRM-skewed circulating memory precursors have been identified (Kok et al, JEM, 2020), and perhaps VV-OVA exposed skin more readily recruits this subset compared to DNFB-exposed skin. Therefore, when the DNFB challenge is performed 5 days after vaccinia virus, the DNFB site may already be at a disadvantage in the recruitment of CD8+ T cells that can efficiently form TRM. In addition, CD8+ T cell-extrinsic mechanisms may be at play, such as differences in myeloid cell recruitment and differentiation or local cytokine and chemokine levels in VV-infected and DNFB-treated skin that could account for differences seen in TRM phenotype and function between these two sites. Although the authors do show that providing exogenous peptide antigen at the DNFB-site rescues their phenotype in relation to the VV-OVA site, the potential antigen-independent factors distinguishing these two sites remain unaddressed. In addition, there is a possibility that peptide treatment of DNFB-treated initiates a second phase of priming of new circulatory effectors in the local-draining lymph nodes that are then recruited to form TRM at the DFNB-site, and that the effect does not solely rely on TRM precursors at the DNFB-treated skin site at the time of peptide treatment.

Thank you for pointing out these potential caveats to our work. We have considered the possibility that late application of peptide or cell-extrinsic difference could affect the interpretation of our results. We would like to highlight that in our prior publication on this topic [1], we found that OT-1 responses in mice infected with VV-OVA and VV-N (irrelevant antigen) yielded the same responses as in our VV-OVA/DNFB models. In addition, in both our prior publication and our current manuscript, application of peptide to DNFB painted sites results in TRM with a similar phenotype to those in the VV-OVA site. Thus, we are confident that it is the presence of cognate antigen in the skin that drives the augmented TRM fitness that we observe.

Secondly, although the authors conclusively demonstrate that TGFBRIII is induced by TCR signals and required for conferring increased fitness to local-antigen-experienced CD8+ TRM compared to local antigen-inexperienced cells, this is done in only one experiment, albeit repeated 3 times. The data suggest that antigen encounter during TRM formation induces sustained TGFBRIII expression that persists during the antigen-independent memory phase. It remains unclear why only the antigen encounter in skin, but not already in the draining lymph nodes, induces sustained TGFBRIII expression. Further characterizing the dynamics of TGFBRIII expression on CD8+ T cells during priming in draining lymph nodes and over the course of TRM formation and persistence may shed more light on this question. Probing the role of this mechanism at other sites of TRM formation would also further strengthen their conclusions and enhance the significance of this finding.

This is an intriguing point. We do not understand why expression of TGFbR3 in TRM required antigen encounter in the skin if TRM at all sites clearly have encountered antigen during priming in the LN. We speculate that durable TGFbR3 expression may require antigen encounter in the context of additional cues present in the periphery or only once cells have committed to the TRM lineage. A more detailed characterization of the dynamics of TGFbR3 expression in multiple tissues would be informative and represents a promising future direction for this project. We note that to robustly perform these experiments a reporter mouse would likely be a requirement.

Reviewer #2 (Public review):

Weaknesses:

Overall, the authors' conclusions are well supported, although there are some instances where additional controls, experiments, or clarifications would add rigor. The conclusions regarding skin-localized TCR signaling leading to increased skin CD8+ TRM proliferation in-situ and increased TGFBR3 expression would be strengthened by assessing skin CD8+ TRM proliferation and TGFBR3 expression in models of high versus low avidity topical OVA-peptide exposure.

Thank you for these helpful suggestions. We did not attempt these experiment as we were concerned that given the relatively modest expansion differences observed with the APL that resolving differences in TGFbR3 and BrdU would prove unreliable. However, this is something that we could attempt as we continue working on this project.

The authors could further increase the novelty of the paper by exploring whether TGFBR3 is regulated at the RNA or protein level. To this end, they could perform analysis of their single-cell RNA sequencing data (Figure 1), comparing Tgfbr3 mRNA in DNFB versus VV-treated skin.

As discussed above, a more detailed analysis of TGFbR3 regulation is of great interest. These experiments would likely require the creation of additional tools (e.g. a reporter mouse) to provide robust data. However, as suggested, we have re-analyzed our scRNAseq looking for expression of Tgfbr3. Pseudobulk analysis of cells isolated from VV or DNFB sites suggests that Tgfbr3 appears to be elevated in antigen-experienced TRM at steady-state (Author response image 1).

Pseudobulk analysis by average gene expression of Tgfbr3 in cells isolated from either VV or DNFB treated flanks, divided by the average gene expression of Tgfbr3 in naïve CD8 T cells from the same dataset.

For clarity, when discussing antigen exposure throughout the paper, it would be helpful for the authors to be more precise that they are referring to the antigen in the skin rather than in the draining lymph node. A more explicit summary of some of the lab's previous work focused on CD8+ TRM and the role of TGFb would also help readers better contextualize this work within the existing literature on which it builds.

We appreciate this feedback, and we have clarified this in the text.

For rigor, it would be helpful where possible to pair flow cytometry quantification with the existing imaging data.

Thank you for these suggestions. In terms of quantification of number of TRMby flow cytometry, we have previously demonstrated as much as a 36-fold decrease in cell count when compared to numbers directly visualized by immunofluorescence [1]. Thus, for enumeration of TRM we rely primarily on direct IF visualization and use flow cytometry primarily for phenotyping.

Additional controls, namely enumerating TRM in the opposite, untreated flank skin of VV-only-treated mice and the treated flank skin of DNFB-only treated mice, would help contextualize the results seen in dually-treated mice in Figure 2.

Without a source of inflammation (e.g. VV infection of DNFB) we see very few TRMin untreated skin. A representative image is provided (Author response image 2). A single DNFB stimulation does not recruit any CD8+ T cells to the skin without a prior sensitization [2].

Representative images of epidermal whole mounts of VV treated flank skin, and an untreated site from the same mouse isolated on day 50 post infection and stained for CD8a.

In figure legends, we suggest clearly reporting unpaired T tests comparing relevant metrics within VV or DNFB-treated groups (for example, VV-OVA PBS vs VV-OVA FTY720 in Figure 3F).

Thank you for this suggestion. The figure legends have been amended.

Finally, quantifying right and left skin draining lymph node CD8+ T cell numbers would clarify the skin specificity and cell trafficking dynamics of the authors' model.

We quantified the numbers of CD8 T cells in left and right skin draining lymph nodes by flow cytometry in mice at day 50 post VV infection DNFB-pull. We observe similar numbers of cells at both sites (Author response Image 3).

Quantification of total number of CD8+ T cells in left and right inguinal lymph nodes.

Each symbol represents paired data from the same individual animal, and this is representative of 3 separate experiments.

Reviewer #1 (Recommendations for the authors):

(1) Figures 1D and S1C demonstrate that 80-90 % of TRM at both VV and DNFB sites express CD103+. In contrast, the sequencing data suggests the TRM at the VV site has much higher Itgae expression. Also, clusters 3 and 4, which express significantly more Itgae than all other clusters, together comprise only ~30% of CD8+ T cells at the VV-infected skin site. How can these discrepancies between transcript and protein expression be explained?

Thank you for these excellent comments. TRM at both VV and DNFB sites appear to express similarly high levels of CD103 protein in both the OT-I system as we previously published [1] and in a polyclonal system using tetramers. The lower penetrance of Itgae expression in the scRNAseq data we attribute to a lack of sensitivity which is common with this modality. However, the relative increased expression of Itgae in clusters 3 and 4 is interesting and may suggest increased Itgae production/stability. However, in the absence of any effect on protein expression, we chose not to focus on these mRNA differences.

(2) For the experiments in Figure 3D, in order to exclude a contribution from circulating memory cells, FTY720 should have been administered during the duration of, not prior to, the initiation of the recall response. The effect of FTY720 wears off quickly, so the current experimental setting likely allows for circulating cells to enter the skin. This concern is mitigated by the results of anti-Thy1.1 mAb treatment, but documenting the experiment as in Figure D will likely be confusing to readers.

Thank you for this comment. We relied on the literature indicating that the half-life of FTY720 in blood is longer than 6 days [3-5]. However, on reviewing this again, there are other reports suggesting a lower halflife. Thank you for pointing out this potential caveat. As mentioned above, we do not think this affects the interpretation of our data as similar results were obtained with anti-Thy1.1

(3) Similar to what is described in the weaknesses section, the data on TGFBRIII expression is lacking. When is TGFBRIII induced? In the LN during primary activation and it is then sustained by a secondary antigen exposure at the peripheral target tissue site? Or is it only induced in the peripheral tissue, and there is interesting biology to uncover in regard to how it is induced by the TCR only after secondary exposure, etc.?

Thank you for these comments. As discussed above, a more detailed analysis of TGFbR3 regulation is of great interest. These experiments would likely require the creation of additional tools (e.g. a reporter mouse) to provide robust data and are part of our future directions.

(4) As described in the weakness section, there could be TCR-independent differences between the VV-OVA and DNFB sites that lead to phenotypic changes in the TRMs that are formed there, both CD8+ T cell-intrinsic (kinetics; with regard to time after initial priming) and extrinsic (microenvironmental differences due to the nature of the challenge, recruited cell types, cytokines, chemokines, etc.). Since the authors report the use of both VV and VV-ova, we recommend an experimental strategy that controls for this by challenging one site with VV and another with VV-OVA concomitantly, followed by repeating the key experiments reported in this manuscript.

As discussed above, we have previously published a very similar experiment using VV-OVA and VV-N infection on opposite flanks [1].

(5) In Figure 6J please indicate means and provide more of the statistics comparing the groups (such as comparing VV-WT vehicle to VV-KO vehicle etc.), and potentially display on a linear scale as with all of the other figures looking at cells/mm2 to help convince the reader of the conclusions and support the secondary findings mentioned in the text such as "Notably, numbers of Tgfbr3ΔCD8 TRM in cohorts treated with vehicle remained at normal levels indicating that loss of TGFβRIII does not affect TRM epidermal residence in the steady state" despite it looking like there is a decrease when looking at the graph.

We appreciate the feedback on the readability of this figure, and so have updated figure 6J to be on a linear scale and added additional helpful statistics to the figure legend. The difference between Tgfbr3WT and Tgfbr3∆CD8 at steady state is excellent point, and we agree that there could to be a trend towards reduction in the huNGFR+ TRM across both groups, even without CWHM12 administration. However, we did not see statistically significant reductions in steady-state Tgfbr3∆CD8 TRM, but the slight reduction in both VV-OVA and DNFB treated flanks suggests that TGFßRIII may play a role in steady-state maintenance of all TRM. Perhaps with more sensitive tools to better visualize TGFßRIII expression, we could identify stepwise upregulation of TGFßRIII depending on TCR signal strength, possibly starting in the lymph node. We have also amended our description of this figure in the text, to allow for the possibility that a low, but under the level of detection amount of TGFßRIII could play a role in steady-state maintenance of both local antigen-experienced and bystander TRM.

Minor points:

(1) In describing Figure 4B, the term "doublets" for pairs of connected dividing cells is confusing.

Thank you for this comment, the term has been revised to “dividing cells” in the text and figure.

(2) Figure legend 4F: BrdU is not "expressed" .

Very true, it has been changed to “incorporation”.

(3) Do CreERT2 and/or huNGFR expressed by transferred OT-I cells act as foreign antigens in C57BL/6 mice, potentially causing elimination of circulating memory cells? If that were the case, this would not necessarily confound the read-out of TRM persistence studied here, since skin TRM are likely protected from at least antibody-mediated deletion and their numbers are not maintained by recruitment of circulating cells at stead-state. However, it would be useful to be aware of this potential limitation of this and similar models.

Thank you for raising the important technical concern. In our prior work [1] and this work, we monitor the levels of transferred OT-I cells in the blood over time. We have not observed rejection of huNGFR+ cells. We also note that others using the same system have also not observed rejection [6].

(4) In Figure 6J, means or medians should be indicated

This has been updated in Figure 6J.

(5) Using the term "antigen-experienced" to specifically refer to TRM at the VV site could be confusing, since those at the DNFB site are also Ag-experienced (in the LN draining the VV skin site).

We agree that it is a challenging term, as all TRM are memory cells. That is why in the text we refer to TRM isolated from the VV site as “local antigen experienced TRM.”, to try to distinguish them from bystanders that did not experience local antigen.

(6) The Title essentially restates what was already reported in the authors' prior study. If the data supporting the TGFBRIII-mediated mechanism is studied in more depth, maybe adding this aspect to the title may be useful?

Thank you for this suggestion. I think the current title is probably most suitable for the current manuscript but we are willing to change it should the editors support an alternative title.

Reviewer #2 (Recommendations for the authors):

(1) Definition of bystander CD8+ TRM: The first paragraph of the introduction defines CD8+ TRM. To improve the clarity of this definition, we suggest being explicit that bystander TRM experience cognate antigen in the SDLNs but, in contrast to other TRM, do not experience cognate antigen in the skin.

Thank you, we have clarified this is in the text.

(2) Consider softening the language when comparing the efficiency of CD8+ recruitment of the skin between DNFB and VV-treated flanks. For example, substitute "equal efficiency" with "comparable efficiency" since it is difficult to directly compare the extent of inflammation between viral and hapten-based treatments.

We have adjusted this terminology throughout the paper.

(3) Throughout figure legends, we appreciate the indication of the number of experimental repeats performed. We suggest, either through statistics or supplemental figures, demonstrating the degree of variability between experiments to aid readers in understanding the reproducibility of results.

Thank you for this suggestion. In key figures we show data from individual mice across multiple experiments. Thus, inter-experiment variability is captured in our figures.

(4) Figure 1:

a) Add control mice treated with either vaccinia virus or DNFB and harvest back skin at day 52 to demonstrate baseline levels of polyclonal and B8R tetramer-positive CD8s in the epidermis. These controls would clarify the background CD8+ expansion that might occur in DNFB-treated mice in the absence of vaccinia virus.

This point was addressed above.

b) Figure 1: It would be helpful to see the %Tet+ population specifically in the CD103+ population, recognizing that the majority of the CD8+ from the skin are CD103+.

We did look only at CD103+ CD8 T cells from the skin for our tetramer analysis, so this has been clarified in the figure legend.

c) Provide a UMAP, very similar to 1H, where CD8+ T cells, vaccinia virus, and DNFB-treated flanks are overlaid.

Thank you for this suggestion. A UMAP combining aspects of 1G (cell types from the whole ImmgenT dataset) with 1H (our data) results in a figure that is very difficult to interpret. Thus, we have separated cell types across the entire ImmgenT data set (e.g. CD8+ T cells) and our data into 2 separate panels.

d) 1D: left flow plot has numbered axis while the right flow plot does not.

Thank you, this has been fixed.

(5) Figure 2:

a) In the figure legend, define what is meant by the grey line present in Figures 2C and 2D.

This has been updated in the figure legend.

b) Edit the Y axis of 2C and 2D to specify the TRM signature score.

This has been updated in the figure.

c) Include panel 1D from 1S into Figure 2 to help clarify for the reader what genes are expressed in the 0 - 5 clusters.

We appreciate the feedback, but we found the heatmap made the figure look too busy, so we feel comfortable keeping it available within supplemental figure 1.

d) In body of text explicitly discuss that the TRM module used to calculate a signature score was created using virus infection modules (HSV, LCMV and influenza) and thus some of the transcriptional similarity between the authors vaccinia virus treated CD8+ TRM and the TRM module might be due to viral infection rather than TRM status.

Thank you for this comment. We have now emphasized this point in the text.

(6) Figure 3:

a) If there are leftover tissue sections, it would be optimal to show specific staining for CD103. We recognize that this data has been previously published by the lab, but it would be ideal to show it once in this paper.

Unfortunately, we do not have leftover tissue sections, so we are unable to measure CD103 by I.F. in these experiments.

b) If you did collect skin draining lymph nodes in the Thy1.1 depletion model, it would be nice to see flow data showing the depletion effects in the skin draining lymph nodes in addition to the blood.

Unfortunately, we did not collect the skin draining lymph nodes, and do not have that data for the relevant experiments.

c) Figure 3 F & G: Perform a T-test comparing vaccinia virus PBS to FTY720 and isotype to anti-Thy1.1 within the same treatment group. Showing no significance with these two comparisons would strengthen the authors' claims. Statistics can be described in legend.

We have included this analysis in the figure legend.

(7) Figure 4:

a) It would be helpful to have the CD69+/CD103+ population in this model discussed/defined more. The CD69 expression seen in 4E is lower than the reviewers would've predicted, and it would be interesting to see CD103 expression as well.

We have found that generally CD103 is a stronger marker for in the skin by flow, as CD69 staining is somewhat less robust in the colors we have chosen. By way of example, we present gating we did upstream in that experiment, gated previously on liveCD45+CD3+CD8+ events (Author response image 4).

Representative flow cytometric plots showing CD69 and CD103 expression in gated live CD45+CD8+CD90.

1+ cells isolates from VV-OVA or DNFB treated flanks.

(8) Figure 5:

a) Define APL and its purpose in both the body of text and the figure legend.

We have clarified this in the text and the figure legend.

b) Using in-vivo BrdU, compare proliferation between high avidity N4 and low avidity Y3 OVA-peptide at the primary recall timepoint.

We considered this, but due to the lack of sensitivity of the BrdU incorporation and the relatively subtle phenotype of the Y3, we did not think the assay would be sensitive enough to identify differences.

(9) Figure 6:

a) Compare TGFBR3 expression in CD8+ T cells from mice receiving high avidity N4 versus low avidity Y3 OVA-peptide at the primary recall timepoint.

This point was discussed above.

b) Either 1) examine TGFBR3 mRNA expression in VV vs DNFB skin from scRNA-seq dataset or 2) perform a qPCR on epidermal CD8+ T cells from mice receiving high avidity N4 versus low avidity Y3 at the primary recall timepoint. This would help distinguish whether TGFBR3 regulation occurs at the mRNA versus protein level.

This point has been discussed above.

c) Figure 6A: Not required, but it seems like the TGFBR3 gate could be shifted to the right a bit.

The gates were set using FMO.

d) Figure 6C: What comparison is the asterisk indicating significance referring to?

It is the Dunnett’s test comparing VV-OVA to DNFB and untreated skin, the figure has been amended to clarify this point.

e) Figure 6: To increase the rigor of the claim that CWHM12 is creating a TGFb limiting condition, the authors could either 1) perform an ELISA or cell-based assay measuring active TGFb, 2) recapitulate results of 6J using monoclonal antibody against avb6 as done in Hirai et al., 2021, Immunity., or 3) examine Tgfbr3 mRNA expression in your single cell RNAseq data, comparing cluster 0 and cluster 3.

We are pleased to have the opportunity to show Tgfbr3 mRNA, which is above in figure R1.

(10) Material and methods:

Specify how the localization of the back skin used for imaging was made consistent between the right and left flanks.

We have updated this methodology in the text.

Literature Cited

(1) Hirai, T., et al., Competition for Active TGFβ Cytokine Allows for Selective Retention of Antigen-Specific Tissue- Resident Memory T Cells in the Epidermal Niche. Immunity, 2021. 54(1): p. 84-98.e5.

(2) Manresa, M.C., Animal Models of Contact Dermatitis: 2,4-Dinitrofluorobenzene-Induced Contact Hypersensitivity, in Animal Models of Allergic Disease: Methods and Protocols, K. Nagamoto-Combs, Editor. 2021, Springer US: New York, NY. p. 87-100.

(3) Müller, H.C., et al., The Sphingosine-1 Phosphate receptor agonist FTY720 dose dependently affected endothelial integrity in vitro and aggravated ventilator-induced lung injury in mice. Pulmonary Pharmacology & Therapeutics, 2011. 24(4): p. 377-385.

(4) Nofer, J.-R., et al., FTY720, a Synthetic Sphingosine 1 Phosphate Analogue, Inhibits Development of Atherosclerosis in Low-Density Lipoprotein Receptor–Deficient Mice. Circulation, 2007. 115(4): p. 501-508.

(5) Brinkmann, V., et al., Fingolimod (FTY720): discovery and development of an oral drug to treat multiple sclerosis. Nat Rev Drug Discov, 2010. 9(11): p. 883-97.

(6) Andrews, L.P., et al., A Cre-driven allele-conditioning line to interrogate CD4+ conventional T cells. Immunity, 2021. 54(10): p. 2209-2217.e6.

Version 2

Abstract

Abstract

CD8+ tissue resident memory T cells (TRM) develop from effectors that seed peripheral tissues where they persist providing defense against subsequent challenges. TRM persistence requires autocrine TGFβ transactivated by integrins expressed on keratinocytes. TRM precursors that encounter antigen in the epidermis during development outcompete bystander TRM for TGFβ resulting in enhanced persistence. ScRNA-seq analysis of epidermal TRM revealed that local antigen experience in the skin resulted in an enhanced differentiation signature in comparison with bystanders. Upon recall, TRM displayed greater proliferation dictated by affinity of antigen experienced during epidermal development. Finally, local antigen experienced TRM differentially expressed TGFβRIII, which increases avidity of the TGFβRI/II receptor complex for TGFβ. Selective ablation of Tgfbr3 reduced local antigen experienced TRM capacity to persist, rendering them phenotypically like bystander TRM. Thus, antigen driven TCR signaling in the epidermis during TRM differentiation results in a lower TGFβ requirement for persistence and increased proliferative capacity that together enhance epidermal TRM fitness.

Introduction

Tissue resident CD8 memory T cells (TRM) are a highly abundant, non-circulating, long-lived subset of memory T cells that play an important role in protecting against re-infections [1, 2]. TRM also provide immunosurveillance against neoplasia and are thought to be pathogenic in some autoimmune diseases [3–9]. In the skin, following infection with vaccinia virus (VV) or herpes simplex virus, TRM develop from CD8+ T cell effectors (TEFF) that expand following priming in the lymph node and are recruited into the skin by inflammatory signals where they preferentially reside in the epidermis [1, 10–15]. Unlike TRM in some tissues, re-encounter with cognate antigen in the skin is not required for TRM differentiation [16–19]. Epidermal TRM that encounter cognate antigen in the skin or bystander TRM that do not encounter antigen both develop with comparable efficiency and both express similar levels of the canonical TRM markers CD103 and CD69[20, 21].

The cytokine transforming growth factor β (TGFβ) is required for cutaneous TRM development at multiple stages. TGFβ signaling in the lymph node during steady state epigenetically preconditions naïve CD8 T cells allowing for later TRM differentiation [22]. TEFF recruited into skin require TGFβ signaling for entry into the epidermis and differentiation into TRM [23, 24]. Once epidermal TRM have differentiated, these cells continue to require TGFβ signaling to retain epidermal residence. TGFβ is produced bound to the latency associate peptide that prevents bioactivity until the complex is activated which in the epidermis is mediated exclusively by activation via the integrins αvβ6 and αvβ8 expressed by keratinocytes [25–28]. Epidermal persistence of TRM depends on autocrine TGFβ which is transactivated by the integrins ⍺vβ6 and ⍺vβ8[21, 29]. Inducible ablation in TRM of a required component of the TGFβ receptor (TGFβRII) or TGFβ1 as well as ablation of ⍺vβ6 and ⍺vβ8 in keratinocytes all result in loss of epidermal TRM [21, 30].

We previously reported that local antigen experienced TRM that have encountered cognate antigen in the skin are better able to persist in the epidermis than bystander TRM that have not encountered antigen in the skin when active TGFβ is limited, despite the fact that both groups of cells had prior activation within the lymph node [21]. This is evident when TGFβ activation is experimentally reduced by small molecule inhibition of ⍺vβ6 and ⍺vβ8 and when established TRM compete with newly recruited TEFF cells for limiting amounts of TGFβ activation. In both instances, bystander TRM are preferentially lost from the epidermis while local antigen experienced TRM persist. Thus, encounter with antigen in the skin results in more fit TRM and represents a potential opportunity to preferentially enrich antigen-specific over bystander TRM cells during repeated challenges.

Herein, we delineate mechanisms of TRM homeostasis by demonstrating that within a population of endogenous TRM, antigen encounter in the skin is required for their full differentiation whereas bystander TRM are maintained at an earlier developmental stage. We also find that local antigen experienced TRM have increased proliferative capacity during a recall response compared to bystander TRM that is dependent on affinity of antigen experienced in skin thus providing an additional functional attribute defining TRM fitness. Finally, we show that expression of TGFβRIII by local antigen experienced TRM which can be induced following TCR ligation is required for epidermal persistence when active TGFβ is limiting.

Results

Epidermal TRM are transcriptionally heterogeneous

The recruitment of effector CD8+ T cells into mouse flank skin via a viral infection (e.g. Vaccinia Virus) or through an inflammatory stimulus (e.g. “DNFB-pull”) results in comparable numbers of long-lived CD103+ epidermal TRM and is independent of the presence or absence of cognate antigen in the skin [18, 29, 31]. This, however, has only been demonstrated using HSV-specific (gBT-I) or OVA-specific (OT-I) TCR transgenic T cells. To determine if endogenous polyclonal CD8+ T cells share this phenotype, we employed a dual VV infection “DNFB-pull” model. Cohorts of wildtype C57BL/6 mice were infected on the left flank with VV by skin scarification (Figure 1A). The infection expanded CD8+ effectors specific for VV antigens which were then recruited into the left flank in response to the ongoing infection-induced inflammation. DNFB is applied to the right flank 5 days post infection to recruit into the skin circulating CD8 effectors expanded by infection. Mice were then rested for 50+ days to allow for the formation of TRM. We have previously found that evaluation of TRM numbers in the epidermis by flow cytometry is much less accurate than direct immunofluorescent visualization [21]. Evaluation of whole mounted epidermal sheets stained with anti-CD8 revealed comparable numbers of CD8+ T cells at the VV infected (left flank) and DNFB treated (right flank) sites (Figure 1B and C). Expression of CD103 on TRM as evaluated by flow cytometry was also similar at both sites (Figure 1D and S1A). To evaluate the frequency of antigen-specific T cells at each site, skin from a separate cohort was examined by flow cytometry using the B8R tetramer which recognizes an immunodominant epitope of VV in H2-Kb[32]. We observed equivalent numbers of B8R+ TRM at the VV-infected and DNFB treated sites (Figure 1E and F). From these data we conclude that polyclonal CD8 T cells form epidermal TRM with comparable efficiency when recruited into the skin by viral infection or sterile inflammation and that the presence or absence of cognate antigen in the skin has no effect on the number or frequency of antigen-specific TRM.

Epidermal TRM are transcriptionally heterogeneous.

(A) Experimental design. Mice were treated with vaccinia virus infection by skin scarification on the left flank on day 0, and then on day 5 post infection the right flank was painted with 0.15% DNFB. On day 56, flanks were harvested for either epidermal whole mount, flow cytometry, or cells were sorted for scRNA-seq. This panel was created using BioRender.com. (B) Representative images and (C) quantification of epidermal whole mounts of VV and DNFB treated flanks harvested on day 50 post infection stained for CD8a. (D) Representative flow plots gated on CD45+ CD3+ CD8+ CD90.2+ cells isolated from VV or DNFB treated flanks. (E) Representative flow plots and (F) quantification of B8R tetramer binding of CD103+ TRM gated as in (D) isolated from VV or DNFB treated flanks. (G) The integrated Minimal-Distorted Embedding of all 96 experiments from the ImmgenT consortium with annotated clusters. (H) Minimal-Distorted embedding visualization of transcriptional clusters of skin TRM projected over the ImmgenT dataset. (I) The percentage of cells in each transcriptional cluster found in VV or DNFB treated sites. Each symbol represents paired data from the same individual animal (C, F). Data shown to be nonsignificant by paired Student’s t-tests (C,F). Data are representative of 3 separate experiments. Scale bar in (B) represents 50um.

Although TRM efficiently populate the epidermis independently of cutaneous cognate antigen, we have previously demonstrated that TRM that form at skin sites containing cognate antigen are functionally distinct from bystander TRM that form in its absence[21]. To analyze potential transcriptional differences between local antigen experienced and bystander TRM, we performed single-cell RNAseq in collaboration with the ImmgenT consortium [33]. The consortium consists of multiple groups that isolated populations of murine T cells from multiple tissues and contexts and then subjected them to single-cell RNA-seq [33](see Methods). 627,692 mature T cells from 80 different experiments and 703 samples were integrated and batch-corrected using totalVI [34] (see Methods).

Cells were projected in two dimensions using Minimal-Distorted Embedding (see Methods) revealing the expected transcriptional clustering of distinct types of T cells (Figure 1G). As part of the consortium, we isolated cells from the flanks of rested (>65 days) VV-infected or DNFB treated flank skin by enzymatic digestion. Cells were pooled from 10 mice and purified based on expression of CD90 and CD8 (FigureS1B). As expected, most cells expressed CD69 and CD103 (Figure S1C). Skin cells from each group and naïve spleen cells were hash tagged and analyzed by single-cell RNA-seq.

Analysis of transcripts revealed a total of 6 clusters (0-5) which are shown overlayed on the complete ImmgenT cell embedding (Figure 1H, S1D). As expected, most cells clustered with CD8⍺β T cells (Figure 1G, black). There was, however, a high degree of heterogeneity and some, possibly contaminating cells fell outside of this region. A comparison of the relative frequency of cells in each of the clusters (0-5) isolated from the VV or DNFB sites revealed a strong enrichment of cluster 3 in the VV group and of clusters 0 and 2 in the DNFB group (Figure 1I, magenta). This suggested that cluster 3 cells were manifesting a distinct transcriptional state likely induced by antigen encounter in the skin.

Cutaneous antigen is required for complete TRM differentiation

To test the hypothesis that cells in cluster 3 could be TRM that have encountered antigen in the skin during development, we performed signature score enrichment analysis comparing the transcriptional profile of our cells to published core genes of well-defined CD8 T cell states. Signature scores were calculated based on normalized differential gene expression compared to core genes of TRM generated by acute infection models (GSE47045), activated T cells (TACT, GSE10239), circulating memory T cells (TMEM, GSE41867), and exhausted T cells (TEX, GS41867) [35–38]. Notably, cells in cluster 3 showed enriched expression of TRM core genes compared to all other clusters (Figure 2A). We performed a similar analysis using pseudo-bulk analysis of all cells isolated from the VV or DNFB sites. Consistent with the enrichment of cluster 3 in cells from the VV site, we found a strong enrichment of TRM core genes in the VV group (Figure 2B). These data suggest that cells in cluster 3 isolated from the VV site represent fully differentiated TRM.

Cutaneous antigen is required for complete TRM differentiation.

(A) Signature score analysis of skin T cell clusters calculating the enrichment for previously published T cell state gene sets (N=naïve). Signature score = Mean (average upgenes z-scores) - Mean (average downgenes z-scores). *** p < 0.001 by Dunnett’s multiple comparisons of cluster 3 to all other clusters. (B) Pseudobulk analysis comparing signature scores of cells isolated from VV and DNFB sites as in (A). ** p < 0.01 and *** p < 0.001 by paired Student’s t-tests. (C) Signature score analysis of individual clusters or (D) cells isolated from VV and DNFB sites calculating the enrichment for compared to epidermal TRM isolated at the indicated day post infection. ***p < 0.001 by Dunnett’s multiple comparison test of cluster 3 to all other clusters in (C). * p < 0.05, ** p < 0.01, **** p < 0.0001 by paired Student’s t-tests in (D). Gray line represents a randomized control generated by the average enrichment of each group compared to a randomly generated gene set of an equal number of probes. Datapoints and error bars represent mean and 95% confidence interval of the relative enrichment of each set of DEGs.

To further test the hypothesis that cluster 3 represents fully differentiated TRM, we performed signature score enrichment analysis of clusters 0-5 against a dataset of skin T cells isolated at different times following skin VV infection (GSE79805) [39]. Cells from clusters 0, 2 and 3 showed similar signature scores in comparison with skin T cells up to day 5 post infection, however clusters 0 and 2 then plateaued suggesting a lack of continued differentiation (Figure 2C). In contrast, cluster 3 transcripts scored higher at later time points in TRM differentiation. Signature scores for cells isolated from the DNFB and VV sites were similar at early time points but diverged at later time points with increased scores observed in cells from the VV site (Figure 2D). Taken together, these data suggest that TRM isolated from skin sites where they can encounter cognate antigen are transcriptionally more fully differentiated TRM while bystander TRM isolated from a site lacking cognate antigen remain in a less differentiated state.

Local antigen experienced TRM have improved expansion in a recall response

Analysis of the 10 highest differentially expressed genes (DEGs) in cluster 3 (Figure S1D) showed increased expression of genes associated with T cell activation including Dusp1 and Nr4a1 as well as the AP-1 family members Junb and Fos [40–44]. Given the importance of AP-1 family members in T cell proliferation and differentiation, we next tested whether this corresponded to functional differences in TRM activity [43, 45–48].

An important functional property of TRM is their capacity to rapidly expand following re-encounter with cognate antigen [49, 50]. To determine whether local antigen experienced and bystander TRM have different recall responses, we repeated our dual VV infection and DNFB “pull” model in combination with an OT-I adoptive transfer model to allow for antigen recall using the SIINFEKL peptide. CD90.1+ OT-I cells were adoptively transferred into naïve C57BL/6 mice followed by infection with a vaccinia virus expressing SIINFEKL peptide, OVA257-264 (VV-OVA) on the left flank. On day 5 post infection, the right flank was treated with DNFB to recruit expanded OT-I effectors to the site (Figure 3A). On day 50+ after infection, topical SIINFEKL peptide or control PBS was painted on both flanks in a single application (1° recall). The total number of OT-I cells in the epidermis 6 days later was determined by immunofluorescent microscopic evaluation of the CD90.1 congenic marker in epidermal whole mounts (Figure 3B-C). Restimulation with peptide led to an expansion of TRM at both sites compared to PBS treated controls. Notably, local antigen experienced OT-I TRM cells in the VV-OVA treated flanks expanded to a larger extent compared to bystander TRM in DNFB treated flanks. To inhibit potential recruitment of circulating effector OT-I cells, we repeated these experiments administering FTY720 to block migration of T cells from lymph node or titrated anti-Thy1.1 antibody to ablate circulating OT-I but not TRM (Figure 3D). Both methods successfully depleted OT-I from the blood (Figure S2A-D), without affecting TRM in the skin (Figure S2E). Administration of FTY720 (Figure 3E-F) or anti-Thy1.1 treatment (Figure 3G) had no effect on the number of OT-I TRM in the epidermis after a 1° recall response and local antigen experienced TRM from the VV-OVA treated flank still expanded to a greater degree than bystander TRM from the DNFB treated flank. Thus, the increased number of epidermal OT-I at the local antigen experienced VV site during a 1° recall does not appear to result from recruitment of circulating cells.

Local antigen experienced TRM have improved expansion in a recall response.

(A) Experimental design. Mice were adoptively transferred with Thy1.1+ OT-I T cells on day −1, then infected with OVA-expressing vaccinia virus by scarification on the left flank on day 0. On day 5 post-infection the right flank was painted with 0.15% DNFB. On day 50, a primary recall response with SIINFEKL peptide in acetone and olive oil was painted on both flanks and harvested on day 56. In some cohorts, mice were allowed to rest for an additional 120 days and then treated with topical SIINFEKL peptide again on day 176 and harvested on day 182. (B) Representative images and (C) quantification of epidermal whole mounts isolated from VV-OVA and DNFB treated flanks stained for anti-Thy1.1 (n=10 animals). (D) Experimental design. Mice were treated as in (A), but for 6 days prior to SIINFEKL treatment they were given either i.p. FTY720, i.p. PBS, i.v. titrated anti-Thy1.1 or i.v. isotype control. (E) Representative epidermal whole mounts images and (F) quantification of skin from VV-OVA treated flanks on day 6 of a primary recall response treated with either FTY720 or PBS. (G) Quantification of total Thy1.1 OT-I cells in epidermal whole mounts on day 6 after a primary recall response in mice treated with either isotype or anti-Thy1.1 depleting antibody. Each symbol represents paired data from an individual same animal (F, G). Data are representative of 3 independent experiments. *p<0.05 by paired Student’s t-tests. Unpaired Student’s t-tests between FTY720 treated (F) or Anti-Thy1.1 treated (G) flanks and PBS treated flanks shows non-signficance for both VV-OVA and DNFB, Scale bar represents 50um (B, F). Each symbol represents the mean +/- SEM (C). Panels A and D were created using BioRender.com.

Finally, to determine whether increased expansion by local antigen experienced TRM persists with multiple rounds of stimulation, we repeated this experiment but allowed mice to rest for 120 days after the 1° recall response. The number of TRM at both the VV-OVA and DNFB sites contracted down to equivalent numbers that were increased compared to PBS treated, control mice (Figure 3C). Mice were then given a 2° recall by a second application of peptide at day +176. The number of TRM expanded but TRM at the VV-OVA site increased by a larger amount than TRM at the DNFB site (Figure 3B-C). Based on these data, we conclude that TRM encountering antigen in the skin during development expand more efficiently in response to antigen re-encounter and this phenotype persists long-term with subsequent antigen encounters.

Local antigen experienced TRM exhibit increased proliferation during a recall response We next hypothesized that the increased expansion of local antigen experienced TRM during a recall response resulted from increased proliferation. To test this, as above, WT mice adoptively transferred with OT-I cells were infected with VV-OVA on the left flank and DNFB on the right flank 5 days later. After at least 50 days of rest, both sides were painted with topical SIINFEKL peptide and mice were administered 2 mg of BrdU i.p. daily (Figure 4A). Epidermal sheets for immunofluorescent visualization were harvested 2 days later. We noted a clear increase in the percent and total number of dividing cells as well as the total number of OT-I T cells in the epidermis at the VV-OVA site compared with the DNFB site (Figure 4B-D). Similar results were obtained by flow cytometry, with an increase in the percentage of BrdU+ in TRM isolated from the VV-OVA site (Figure 4E-G). Staining for Annexin V was equivalent in both groups suggesting that an altered rate of apoptosis does not contribute to the observed changes in expansion (Figure 4H-I). In sum, local antigen experienced TRM show augmented proliferation during an antigen recall response compared with bystander TRM. Moreover, an enhanced proliferative capacity following antigen restimulation can now be added to other parameters of TRM fitness including enhanced epidermal persistence when active TGFβ is limited.

Local antigen experienced TRM have increased proliferation during a recall response.

(A) Experimental scheme. This panel was created using BioRender.com. (B) Representative images of epidermal whole mounts of VV-OVA or DNFB treated flanks on day 2 of a recall response. Arrows highlight cell doublets. (C) Quantification showing percent OT-I cells that are dividing in epidermal whole mounts on day 2 after primary recall response, and (D) total number of dividing OT-I cells. (E) Representative flow cytometric plots and (F) quantification showing BrdU incorporation in gated CD45+ CD8+ CD90.1+ CD103+ CD69+ cells isolates from VV-OVA or DNFB treated flanks. (G) Quantification of total numbers BrdU+ OT-I cells combining OT-I numbers with percentage of BrdU incorporation in (F). (H) Representative histograms and (I) quantification of Annexin V expression in OT-I cells isolated from VV-OVA or DNFB treated flanks or OT-I cells heat-treated at 60°C for 1 hour (HK). Data are representative of 3 separate experiments. *p<0.05, **p<0.01 by Students paired t-tests (C), (D), (F) and (G) or Dunnett’s test (I). Scale bar represents 50um.

TRM fitness depends on TCR signal-strength

We next hypothesized that the strength of TCR signal provided in the skin during TRM differentiation should correlate with TRM phenotype at late time points. To generate TRM with a range of local antigen encounter signal strengths, we wanted to expose newly-recruited T cells to either topical SIINFEKL peptide (N4), or a topical altered peptide ligand (APL) SIYNFEKL (Y3) that has ¼ the avidity for the TCR receptor [51]. We hypothesized that encountering an APL would induce an intermediate functional phenotype between full-strength SIINFEKL and no local antigen encounter. To test this hypothesis, WT mice were adoptively transferred with OT-I cells then infected with VV-OVA on the left flank and 5 days later were challenged with DNFB on the right flank (Figure 5A).One day later, the DNFB site was painted with PBS, SIINFEKL (N4), or an APL SIYNFEKL (Y3), After 50+ days of rest, all mice were challenged in a 1° recall response with topical N4 or control PBS at the DNFB sides (Figure 5A). Epidermal sheets for immunofluorescence were harvested 6 days later. As expected, mice restimulated with PBS showed equivalent numbers of TRM regardless of which peptide was provided during development (Figure 5B-C). Following a 1° recall response, TRM exposed to N4 (DNFB + N4) expanded to a level equivalent to TRM from the VV-OVA site. In contrast, TRM exposed to Y3 (DNFB + Y3) showed a degree of expansion that was intermediate compared to TRM that did not experience antigen in the skin (DNFB+PBS) and those that experienced SIINFEKL (DNFB + N4). Thus, the strength of TCR engagement determines the fitness of TRM based on their capacity to expand during a 1° recall response.

TRM fitness depends on TCR signal-strength.

(A) Experimental scheme. Local antigen experienced and bystander TRM were generated and restimulated as previously described, but on the DNFB-treated flanks, altered peptide ligand variants of SIINFEKL were topically applied to the skin 1 day after recruitment to the skin. This panel was created using BioRender.com. (B) Representative images and (C) quantification of epidermal whole mounts (CD90.1 cyan) of VV-OVA and DNFB+ APL treated flanks, at steady state (no recall) or 6 days post recall response. Each symbol represents data from an individual animal. Data are representative of 5 independent experiments. *p<0.05 by unpaired Student’s t-tests. Scale bar, 50um.

Local antigen experienced TRM have improved persistence mediated by TGFßRIII

A feature of local antigen experienced TRM is that they are better able to persist in the epidermis than bystanders when levels of activated TGFβ are limited, either artificially or during competition with newly recruited TEFF [21]. We have previously found that expression of the canonical TGFβ receptors, Tgfbr1 and Tgfbr2 have equivalent expression in local antigen experienced and bystander TRM [21]. However, there is a third TGFβ receptor, TGFβRIII, that lacks a signaling component but functions as a reservoir of ligand for TGFβ, increasing avidity of the TGFβ receptor [52, 53]. Notably, expression of TGFβRIII has been reported to increase in T cells following TCR ligation [54]. This was confirmed in vitro with anti-CD3 anti-CD28 stimulated OT-I splenocytes (Figure 6A). We also observed increased expression of TGFβRIII in vivo comparing local antigen experienced OT-I TRM with bystanders (Figure 6B-C). Based on this data, we hypothesized that increased expression of TGFβRIII on antigen experienced TRM could explain their capacity to maintain epidermal residence when available TGFβ is limiting.

Local antigen experienced TRM have improved persistence mediated by TGFßRIII.

(A) Representative flow cytometric plots of TGFßRIII staining of Thy1.1+ OT-I cells stimulated in vitro for 48 hours with PBS or anti-CD3 anti-CD28. (B) Representative histograms showing TGFßRIII expressing in CD45+ CD3+ CD8+ CD90.1+ gated OT-I cells isolated from VV-OVA, DNFB or untreated flanks at least 50 days post infection. (C) Quantification of (B). (D) Schematic demonstrating genetics of Tgfbr3WT and Tgfbr3ΔCD8 mice. (E) Representative histogram and (F) quantification of TGFßRIII expression in huNGFR- or huNGFR+ Tgfbr3ΔCD8 T cells harvested 5 days following i.p. treatment with tamoxifen then stimulated in vitro for 48 hours with anti-CD3 and anti-CD28. (G) Experimental scheme. (H) Representative histogram of CD45.2+CD3+CD8+CD90.1+ OT-I cells isolated from LNs after tamoxifen treatment demonstrating transformation efficiency. (I) Representative epidermal whole mounts showing Thy1.1 staining (green), huNGFR staining (red) or merge (yellow) of VV-OVA or DNFB treated flanks from mice adoptively transferred with either Tgfbr3WT or Tgfbr3ΔCD8 cells, treated with tamoxifen i.p., and then given either PBS or i.p. CWHM12 for 10 days. Hair follicles in the sample present as long yellow streaks. (J) Quantification of total huNGFR+ Thy1.1+ OT-I in (I). Each symbol represents data from an individual animal. Black lines represent group means. Data is representative of 3 independent experiments. *p<0.05 by Dunnett’s test (C) or paired Student’s t-tests (F) and (J). Unpaired Student’s t-tests between Tgfbr3WT and Tgfbr3ΔCD8 vehicle treated groups show non-significance for both VV-OVA and DNFB treated flanks. Scale bar represents 50um. Panels D and G were created using BioRender.com.

To test this, we generated OT-I Thy1.1+ E8i-creERT2 huNGFR Tgfbr3fl/fl (Tgfbr3ΔCD8) mice which allow for inducible ablation of Tgfbr3 from TRM as well control OT-I Thy1.1+ E8i-creERT2 huNGFR Tgfbr3WT/WT (Tgfbr3WT) mice (Figure 6D). To validate these mice, Tgfbr3ΔCD8 mice were treated with 0.05mg/g tamoxifen i.p. for 5 days. CD8+ T cells from tamoxifen treated Tgfbr3ΔCD8 mice were then isolated from spleen and lymph nodes and cultured for 2 days with anti-CD3 anti-CD28. Cells were evaluated by flow cytometry for expression TGFβRIII and huNGFR, an indicator of successful induction of cre-mediated excision by tamoxifen. In the huNGFR-negative population where cre was not activated, approximately 50% of the cells expressed TGFβRIII (Figure 6E-F). In contrast, TGFβRIII was largely absent from cells co-expressing huNGFR indicative of efficient ablation of Tgfbr3.

To ablate TGFβRIII from TRM once they have fully differentiated in the epidermis, we next adoptively transferred CD8+ T cells isolated from either Tgfbr3WT or Tgfbr3ΔCD8 mice into naïve WT mice followed by skin VV-OVA infection on the left flank. On day 5 post infection, the right flank was treated with DNFB. After at least 50 days of rest to allow for TRM formation, both cohorts were treated daily with 0.05mg/g tamoxifen i.p. to transform approximately 50% of adoptively transferred cells, generating populations of both control huNGFR negative and huNGFR positive OT-I cells. To test the effects of TGFβRIII ablation, TGFβ activation was inhibited for 10 days by i.p. administration of CWHM12, a small molecule inhibitor of integrins ⍺vβ6 and ⍺vβ8 (Figure 6G-H).

Analysis of epidermal whole mounts by immunofluorescence revealed that blockade of TGFβ activation reduced the number of bystander Tgfbr3WT TRM at the DNFB site, but not local antigen experienced TRM at the VV-OVA site, consistent with our earlier results (Figure 6I-J). In contrast, local antigen experienced huNGFR+ Tgfbr3ΔCD8 TRM from the VV-OVA site that lack TGFβRIII were efficiently depleted from the epidermis by CWHM12 treatment. The depletion of bystander huNGFR+ Tgfbr3ΔCD8 TRM at the DNFB site was augmented. Notably, numbers of Tgfbr3ΔCD8 TRM in cohorts treated with vehicle did not induce a statistically signficant reduction in steady-state TRM, indicating that loss of TGFβRIII is not an absolute requirement for TRM epidermal residence in the steady state. Thus, expression of TGFβRIII by local antigen experienced TRM which is downstream of TCR ligation is required for their capacity to remain in the epidermis when activated TGFβ is limiting.

Discussion

Herein we have demonstrated that the increased capacity of local antigen experienced TRM to persist in the epidermis when levels of TGFβ are limited is mediated by increased expression of TGFβRIII. We also show that local antigen experienced TRM have increased proliferative capacity during repeated antigen recalls. In addition, the increased proliferative capacity was directly correlated with the strength of TCR stimulation during TRM development. Finally, we found that local antigen experienced TRM appear more transcriptionally related to fully differentiated TRM. Taken together, these data support a model in which TCR engagement by cognate antigen in the skin is a required final step in TRM differentiation resulting in their increased fitness exemplified by increased proliferative capacity and the ability to persist in the epidermis when active TGFβ is limited.

We propose that the augmented fitness of local antigen experienced TRM represents a mechanism to enrich for high avidity TCR-clones in the epidermis. Skin inflammation recruits TEFF into the skin, some of which develop into TRM. In the absence of competition with pre-existing TRM, TRM form comparably in the presence or absence of cognate antigen. Thus, we found equivalent numbers of bystander TRM at the DNFB and local antigen experienced TRM at the VV sites with both TCR transgenic and endogenous T cells. In contrast, when new TEFF are recruited into sites with pre-existing TRM, there is clonal competition for limited amounts of active TGFβ resulting in enrichment of fitter, local antigen experienced TRM [21]. We now find that this enrichment likely results from 2 different competitive advantages. First, antigen encounter in the skin results in increased expression of TGFβRIII which increases TGFβ avidity for the signaling by the TGFβ receptor. Since TGFβ signaling is required for epidermal persistence, this would provide an advantage for local antigen experienced TRM over bystanders. Second, local antigen experienced TRM have increased proliferation when re-encountering antigen in the epidermis. Following repeated challenges which would be expected outside of SPF conditions, the combination of improved expansion and persistence would work together to enrich for high avidity clones thereby shaping the epidermal CD8+ T cell memory pool. Recently, it has been observed that TRM can contribute significantly to the pool of circulating memory cells [55–59]. Thus, mechanisms augmenting epidermal TRM fitness that shape the pool of epidermal TRM may also affect the pool of systemic memory cells and represent an example of extra-thymic clonal section.

When TRM were challenged in a primary antigen recall response, we noted that local antigen experienced TRM expanded to a greater extent than bystanders. This expansion resulted from increased in-situ proliferation with minimal contribution from newly recruited TEFF, consistent with prior reports [20, 60, 61]. Interestingly, a similar phenomenon occurred following a second encounter with antigen. This indicates that encounter with peptide at a late time point after TRM differentiation (>50 days) is insufficient to convert bystander TRM into local antigen experienced TRM. Thus, there appears to be a window during TRM development when TCR engagement can allow for full differentiation. We also observed after a single recall response that TRM contracted to an elevated baseline suggesting an increase in the epidermal niche. We speculate this may result from a reduced T cell intrinsic requirement for survival and/or homeostatic proliferation factors such as IL-7 or IL-15 or increased expression of these factors by keratinocytes[62, 63]. Altered sensitivity or availability of TGFβ is unlikely to explain the increased niche size as this would be predicted to vary between local antigen experienced and bystander.

Transcriptional analysis of TRM isolated from the small intestine have revealed intra-organ heterogeneity, with unique transcriptional populations arising early during TRM development [64–66]. This aligns well with our identification of 6 distinct transcriptional clusters of epidermal TRM. Cluster 3 appears to represent fully differentiated TRM based on comparison with other TRM datasets. In addition, cluster 3 cells more highly expressed the activation and proliferation associated genes Junb, Fos and Dusp1 as well as Nr4a1. Increased basal expression of the AP-1 family members Junb and Fos could contribute to the enhanced proliferation of antigen-experienced epidermal TRM during a recall response. Intriguingly, memory CD8 T cells lacking the transcription factor Zbtb20 manifest elevated expression of AP-1 family members and mount more robust antitumor responses [71]. The Nr4a1 gene encodes for Nur77, which is induced by TCR signaling and its expression correlates with peptide avidity. Notably Nur77 is required for TRM formation in the liver [23–25, 67, 68]. Interestingly, cells in cluster 3 only accounted for 27% of TRM that had the opportunity to encounter their cognate antigen in the VV treated flank. We speculate that not all clones at the VV site fully develop into fitter TRM due to lower TCR avidity or specificity to viral antigens only expressed early during infection which would be absent once the clones arrived into skin.

In sum, TCR signaling during TRM differentiation represents a previously unappreciated final step in TRM differentiation. This results in fitter TRM with a lower requirement for TGFβ transactivation due to increased expression of TGFβRIII and enhanced proliferation in response to peptide stimulation. Moreover, the differing responses to altered peptide ligands indicate that the degree of fitness depends on TCR signal-strength. Thus, polyclonal TRM likely develop into a spectrum of bystander to local antigen experienced cells based on TCR avidity. Though we have focused entirely on epidermal T cells, we suspect that these mechanisms may play a role in other epithelial tissues where residency is also dependent upon TGFβ. Additionally, we have solely investigated memory CD8+ T cells after acute inflammation, the role of ongoing TCR-engagement during chronic antigen encounter remains unexplored.

Materials and Methods

Mice

We generated Thy1.1+Rag−/−OT-I mice by crossing OT-I mice with Rag−/− and Thy1.1 mice. E8I-creERT2 and ROSA26.LSL.hNGFR reporter mice were developed by Dario A.A. Vignali (University of Pittsburgh) [29]. E8I-creERT2 mice and Thy1.1+Rag-/- OT-I mice were crossed with ROSA26.LSL.hNGFR reporter mice to generate Tgfbr3WT mice and additionally with Tgfbr3loxP mice to obtain Tgfbr3ΔCD8 mice. Tgfbr3loxp mice were developed by Herbert Y Lin (Program in Membrane Biology/Nephrology, Department of Medicine, Massachusetts General Hospital and Harvard Medical School, Boston, Massachusetts) [69]. We used age- and sex-matched female mice that were between 6 and 12 weeks of age at the start of all experiments. All mice were maintained under specific-pathogen-free conditions and all animal experiments were approved by University of Pittsburgh Institutional Animal Care and Use Committee.

TRM cell models and blocking TGFβ activation treatments

Mice were infected by skin scarification (skin infection) with 3 × 106 plaque-forming units recombinant vaccinia virus expressing the SIINFEKL peptide of ovalbumin (VV-OVA), or vaccinia virus without recombinant peptide expression (VV). For skin scarification, 45 μL of VV-OVA or VV was applied to shaved left flank (4–5 cm2) and the skin were gently scratched 100 times with 27 G needle under anesthesia. VV or VV-OVA infected mice were further treated with 0.15% 1-Fluoro-2,4-dinitrobenzene (DNFB, D1529: Sigma-Aldrich) in 4:1 acetone: olive oil (O1514, Sigma-Aldrich) on the flank opposite the site of infection (40 μL) 5 days post infection. In some experiments, 4:1 acetone and olive oil was added to the stock solution of OVA257–264 (SIINFEKL or SIYNFEKL, RP101611, Genscript/Fisher) in DMSO (D2650, Sigma-Aldrich) (10 mg/mL) for a final concentration of OVA257–264 at 0.2 mg/mL and 50 μL was painted to the DNFB-treated skin 1 day after DNFB treatment.

Recall response

50+ days after VV-OVA infection. 4:1 acetone and olive oil was added to the stock solution of OVA257–264 (SIINFEKL, RP101611, Genscript/Fisher) in DMSO (D2650, Sigma-Aldrich) (10 mg/mL) for a final concentration of OVA257–264 at 0.2 mg/mL and 50 μL was painted to the both flanks.

CWHM12 treatments

For blocking TGFβ activation with a small molecule integrin inhibitor, compound CWHM12 (kindly provided by Indalo therapeutics, Cambridge, MA) was solubilized in 100% DMSO. Further dilution to 50% DMSO was made in sterile 100% PBS and dosed to 100 mg per kg body weight per day delivered by i.p. injections once a day for 10 days.

Adoptive transfers

OT-I cells were purified from spleen and lymph nodes by MojoSort Mouse CD8 T Cell Isolation Kit (48007, Biolegend) according to the manufacturer’s instructions and 1 × 105 OT-I cells were intravenously transferred is all experiments. All mice were allowed to rest for 1 day before further experimentation.

Tamoxifen treatment

Tamoxifen (T5648; Sigma-Aldrich) was dissolved in 1/10th volume of 200 proof ethanol following incubation at 37 °C for 15–30 min with 300 rpm shaking. Corn oil (C8267, Sigma-Aldrich) was added for a final concentration of Tamoxifen at 10 mg/ml and was administered to mice for 5 consecutive days by intraperitoneal injection at 0.05 mg per g body weight, in order to transform approximately 50% of adoptively transferred cells.

BrdU treatment

BrdU (B5002, Sigma) was dissolved in sterile PBS. 2 mg in 200uL of PBS was injected i.p. daily into each mouse for 2 consecutive days before harvest.

Immunofluorescence of epidermis

Epidermal sheets were prepared as previously described [30]. Briefly, 2 square cm sections were harvested 2 cm distal from the mid-clavicular line for consistency, and then the epidermal side of shaved defatted flank skin was affixed to slides with double-sided adhesive (3M, St. Paul, MN). Slides were incubated in 10 mM EDTA in PBS for 90 min at 37°C, followed by physical removal of the dermis. The epidermal sheets were fixed in 4% PFA at RT for 15 min. The epidermal sheets were blocked with PBS containing 0.1% tween-20, 2% BSA and 2% rat serum for 1 hour at RT before staining 1 hour with antibodies at RT in PBS containing 0.1% tween-20 and 0.5% BSA. The slides were mounted with ProLong™ Gold Antifade Mount with DNA stains DAPI (P36931, Thermofisher). Images were captured on Keyence BZX800 fluorescent microscope (Keyence, Osaka, Japan). Analysis was performed using BZ-H4A Advanced Analysis Software(Keyence, Osaka, Japan). For enumeration of cells, three images from distant sites randomly determined within an epidermal sheet from a mouse were counted (total 3 mm2) manually after image processing by Adobe Photoshop (version 6) and the average number per mm2 epidermis was calculated as representative of the epidermal sheet. Anti-CD8α (53–6.7), Thy1.1 (OX-7) and huNGFR (ME20.4), were purchased from Biolegend.

Depletion of circulation cells

For depletion of circulating memory OT-I cells, OT-I adoptive transfer VV-OVA and DNFB treated mice were injected i.p. with 0.3–1 μg anti-Thy1.1 (HIS51, ThermoFisher) in 200 μL PBS for 10 consecutive days. Depletion of OT-I T cells (< 0.5% of total CD8+ T cells and < 1%) were confirmed by staining with anti-Thy1.1 (OX-7) using blood 3 days after depletion. For FTY720 treatment, OT-I adoptive transfer VV-OVA and DNFB treated mice were injected i.p. with FTY720 (10006292, Fisher/Cayan Chemical) at 1 μg/g body weight in sterile PBS with 0.5% DMSO (D2650, Sigma-Aldrich) daily for 10 days. Depletion of circulating cells was confirmed with staining of anti-Thy1.1 (OX-7) and anti-CD8α (53–6.7) using blood 3 days after the end of treatment.

Flow cytometry

Preparation of single cells suspension from skin, shaved skin was harvested, and fat tissues was removed by forceps mechanically. The skin was then mechanically digested with either scissors or a gentleMACS Dissociator (Miltenyi Biotec, Bergisch Gladbach, Germany), and then resuspended in RPMI 1640 media (Gibco, Grand Island, NY) containing 2.5 mg/ml Collagenase XI (9001-12-1, Sigma-Aldrich), 0.1 mg/ml DNase (04536282001, Sigma-Aldrich), 0.01 M 4-(2-hydroxyethyl)-1-piperazineethanesulfonic acid (7365-45-9, Sigma-Aldrich), and 10% fetal bovine serum, followed by incubation in a shaking incubator for 30 minutes at 37 °C. The resulting cell mesh was filtered through a 40-μm cell strainer (BD Biosciences, San Jose, CA). The single-cell suspension was then resuspended in 44% percoll (89428-524, VWR) in RPMI and layered over 67% percoll in PBS. After 20 minutes of centrifugation at 931 x g at RT, the interface was isolated and resuspended in FACs buffer. LNs (axillary and inguinal) were incubated in 400 U/ml Collagenase D (Roche Applied Science, Penzberg, Germany) and 0.1 mg/ml DNase in RPMI 1640 with 10% fetal bovine serum for 40 minutes at 37 °C and then minced through a 40-μm cell strainer. In some experiments LNs were heat-treated at 60°C in a hot water bath for 1 hour. Blood was collected with heparin (H3393, Sigma-Aldrich) and treated with red blood cell lysing buffer (R7757, Sigma-Aldrich). Single-cell suspensions were blocked with 2.4G2 culture supernatant (ATCC, Manassas, VA). Surface staining was performed in standard FACS buffer for 30 minutes at 4 °C. Anti-CD8α (53–6.7), CD44 (IM7), CD3 (17A2), Thy1.2 (30-H12), Thy1.1 (OX-7), CD69 (H1.2F3), CD103 (2E7), BrdU (3D4), MHC-II (M5/114.15.2), CD45.2 (104) and Annexin V were purchased from Biolegend. Anti-TGFßRIII (1C5H11) was purchased from Sigma-Aldrich. Soluble tetrameric B8R20–27/H2-Kb complex (made by the NIH Tetramer Core Facility) was conjugated to PE-labeled streptavidin. A BD LSRFORTESSA (BD Biosciences) and Flowjo software (TreeStar, Ashland, OR) were used for analysis.

Single-cell RNA and Totalseq C-Sequencing - Sample Preparation

Flow cytometry and HT antibody staining

Single-cell lymphocyte suspensions were isolated as described above and stained with anti-CD45-APC, anti-CD8a, anti-CD44, anti-CD103 and different TotalSeq-C Anti-Mouse Hashtags (TotalSeq-C Anti-Mouse Hashtags 1-10 Biolegend #155861-155879, in staining buffer (DMEM, 2% FCS and 10mM HEPES)) for 20min at 4°C in the dark. Standard spleen cell suspensions were stained under identical conditions but included anti-CD45-FITC (BioLegend 109806, clone 104) to differentiate them from other samples when pooled. Each sample was labeled with a unique hashtag, enabling the downstream assignment of individual cell (10x cell barcode) to their respective source samples [70].

Cell sorting

For each sample, 500-5000K DAPI- CD45+ Thy1.2+ CD8a+ (skin samples) or DAPI- CD45+ Thy1.2+ CD8a+ CD44+ CD103+ (LN samples) T cells were sorted on the FACSAria III and pooled in a single collection tube at 4C. DAPI was added just before the sort.

ImmGenT TotalSeq-C custom mouse panel staining

The pooled single cell suspension was stained with the ImmGenT TotalSeq-C custom mouse panel, containing 128 antibodies (Biolegend Part no. 900004815) and FcBlock (Bio X Cell #BE0307). Because 500,000 cells are required for staining, unstained splenocytes were spiked in to reach a total of 500,000 cells.

Second sort

Cells were subsequently sorted a second time with the addition of DAPI to select for live T cells, and include 5,000 total splenocyte standard cells were sorted into a single collection tube.

Single-cell RNA and TotalseqC-Sequencing - Library Preparation

Cell encapsulation and cDNA library. Single-cell RNA sequencing was performed using the 10x Genomics 5’ v2 platform with Feature Barcoding for Cell Surface Protein and Immune Receptor Mapping, adhering to the manufacturer’s guidelines (CG000330). After cell encapsulation with the Chromium Controller, reverse transcription and PCR amplification were performed in the emulsion. From the amplified cDNA library, smaller fragments containing TotalSeq-C-derived cDNA were separated for Feature Barcode library construction, while larger fragments containing transcript-derived cDNA were preserved for TCR and Gene Expression library generation. Library sizes for both cDNA fractions were evaluated using the Agilent Bioanalyzer 2100 High Sensitivity DNA assay and quantified with a Qubit dsDNA HS Assay kit on a Qubit 4.0 Fluorometer.

RNA library construction

After enzymatic fragmentation and size selection of the cDNA, the library was ligated to an Illumina R2 sequence and indexed using unique Dual Index TT set A index sequences, SI-TT-B6.

TotalSeq-C library construction

Totalseq-C derived cDNA was processed into the Feature Barcode libraries following the manufacturer’s protocol. The library was indexed with a unique dual index TN set TN set A (10x part no. 3000510) index sequence SI-TN-F9.

Sequencing

The three libraries were pooled based on molarity in the following proportions: 47.5% RNA, 47.5% Feature Barcode, and 5% TCR. The pooled libraries were sequenced on an Illumina NovaSeq S2 platform (100 cycles) using the 10x Genomics specifications: 26 cycles for Read 1, 10 cycles for Index 1, 10 cycles for Index 2, and 90 cycles for Read 2. Single-cell RNA, TCR and TotalseqC-sequencing - Data processing Code. Code is available on https://github.com/immgen/immgen_t_git/

Count matrices

Gene and TotalSeq-C antibody (surface protein panel and hashtags) counts were obtained by aligning reads to the mm10 (GRCm38) mouse genome using the M25 (GRCm38.p6) Gencode annotation and the DNA barcodes for the TotalSeq-C panel. Alignment was performed with CellRanger software (v7.1.0, 10x Genomics) using default parameters. Cells were identified and separated from droplets with high RNA and TotalSeq-C counts by determining inflection points on the total count curve, using the barcodeRanks function from the DropletUtils package.

Sample demultiplexing

Sample demultiplexing was performed using hashtag counts and the HTODemux function from the Seurat package (Seurat v4.1). Doublets (droplets containing two hashtags) were excluded, and cells were assigned to the hashtag with the highest signal, provided it had at least 10 counts and was more than double the signal of the second most abundant hashtag. Hashtag count data were also visualized using t-SNE to ensure clear separation of clusters corresponding to each hashtag. All single cells from the gene count matrix were uniquely matched to a single hashtag, thereby linking them unambiguously to their original sample.

Quality control (QC) and batch correction

Cells meeting any of the following criteria were excluded from the analysis: fewer than 500 RNA counts, dead cells with over 10% of counts mapping to mitochondrial genes, fewer than 500 TotalSeq-C counts, or positivity for two isotype controls (indicating non-specific TotalSeq-C antibody binding). Non-T cells were excluded based on the expression of the MNP gene signature, B cell signature, ILC gene signature, and the absence of T cell gene signature (score calculated using AddModuleScore_UCell). CITEseq data did not meet quality control and was not used in the analysis.

ImmGen T integration

The data was integrated with the rest of the ImmgenT dataset using the SCVI.TOTALVI model (v1.2.0) and the 10x lane as a batch parameter. Dimensionality reduction was performed using the pymde.preserve_neighbors() function with default parameters [CITATION: https://pymde.org/citing/index.html]. Cell clustering was carried out using the FindClusters() function in Seurat. Manual annotation by the immgenT consortium was done using protein and RNA expression of Cd3e, Trbc1, Trbc2, Cd4, Cd8a, Cd8b1, Foxp3, Mki67, Sell, Cd44, Trgc1, Itgax, Itgam, Ms4a1, CD3, TCRB, THY1.2, CD4, CD8A, CD8B, CD62L, CD44, TCRGD, TCRVG1.1, TCRVG2, TCRVG3, CD19, CD20, ITAM.CD11B, ITAX.CD11C, KLRBC-NK1.1. Data was visualized using the Rosetta software (https://cbdm.connect.hms.harvard.edu/ImmgenT/PublicRosetta/). ImmgenT integration and annotation available on https://www.immgen.org/ImmGenT/.

Clustering and dimensionality reduction

Using Seurat v4.2. [https://pubmed.ncbi.nlm.nih.gov/31178118/], the variance-stabilizing transformation (VST method) was applied, and PCA was conducted on the top 2000 genes. Principal components explaining 80% of the total variance were selected for two-dimensional reduction using UMAP. Clustering was performed on these principal components using the FindClusters() function in Seurat.

Differential Expression

To determine differentially expressed genes, FindMarkers() within Seurat was used. AddModuleScore() was used to visualize aggregated expression of a set of genes. Volcano plots were created using EnhancedVolcano (v1.14.0).

Signature score analysis

Signature score was calculated by first generating differentially expressed genes of each cluster or condition compared to naïve T cells from the spleen control, and then calculating an average z-score for every up and down gene in a core gene signature datasets. The core gene datasets were generated by Jaiswal et al. [35] by calculating the exclusive expression of genes from T-cells in the datasets GSE47045 [23], GSE10239 [37] and GSE41867 [38]. Comparative programs for TRM development over time were analyzed against the transcripts taken from the top 100 differentially expressed genes of each TRM timepoint/ naïve from GSE79805 [39] by adapting published scoring methods with this data set Jaiswal et al. [35]. The overall signature score was scored based on expression data that was quantified using RSEM in transcripts per million (TPM), then log-transformed as log2(TPM+1). It was then centered for each gene across all cells. The centered data was averaged across sets of genes to define signature scores. We subtracted a control score from the signature score, which is defined using the same process on randomly selected gene-sets. A randomized control was generated by the average enrichment of each group compared to a randomly generated gene set of an equal number of probes, as iterated Jaiswal et al. [35].

Quantification and statistical analysis

Groups were compared with Prism software (GraphPad) using two-tailed paired Student’s t-test for comparison of left and right flanks, two-tailed unpaired Student’s t-test for comparison between animals, or Dunnett’s test for comparisons of more than two groups. Data is presented as each data point and mean with the standard error of the mean (s.e.m.). p < 0.05 was considered significant.